
## Overview

Section for listing best programming practices in R.

## Styling

### pretty code

according to tidyverse standard. Use [styler](http://styler.r-lib.org/) package

### colorize code in terminal

```{r prettycode-pkg-prettycode-function}
```

## R-settings

`.Rprofile` is used as settings file.

### package availability in interactive sessions

Run the following `usethis` functions for adding snippets to `.Rprofile`

1. `usethis::use_devtools()`: makes devtools available in interactive sessions.
2. `usethis::use_usethis()`: makes usethis available in interactive sessions.
3. `usethis::use_reprex()`: makes reprex available in interactive sessions.
4. `usethis::use_conflicted()`: makes conflicted available in interactive sessions.
5. `usethis::use_partial_warning()`: warns on partial matches.

## Package Development

### Publishing

for easy package websites use [pkgdown](https://pkgdown.r-lib.org/articles/pkgdown.html)

### Development

check out [usethis](https://github.com/r-lib/usethis) package

### Use tidyverse conventions

read the usethis help on [tidyverse development](https://usethis.r-lib.org/reference/tidyverse.html)

## File System handling

`fs` [package](https://fs.r-lib.org/) is designed to handle these issues

```{r fs-pkg-quick-demo}
```

## Namespace Conflicts

Check out `conflicted` [package](https://conflicted.r-lib.org) to handle namespace conflicts

## Factors (Categorical) Data Handling

the `forcats` [package](https://forcats.tidyverse.org/) is developed for handling `factors`, R's categorical data.
Also check out the Factors section \@ref(factors)

```{r forcats-pkg-quick-demo}
```

```{r forcats-pkg-cross-demo}
```

```{r forcats-pkg-lump-min-demo}
```

```{r forcats-pkg-matching-demo}
```

```{r forcats-pkg-as_factor-demo}
```

```{r forcats-pkg-relevel-demo}
```

## Working with data bases

the `dbplyr` [package](https://dbplyr.tidyverse.org/) is designed for this

```{r dbplyr-pkg-quick-demo}
```


## function checks

Given

```{r}
wt_mean <- function(x, w) sum(x * w) / sum(x)
wt_mean(1:6, 1:3)
```

there's gotchas with R's vector recycling rules. Use checks

```{r}
wt_mean <- function(x, w) {
  if (length(x) != length(w)) {
    stop("`x` and `w` must be the same length", call. = FALSE)
  }
  sum(w * x) / sum(x)
}
```

`x` and `w` must be the same length

```{r, eval = FALSE}
wt_mean(1:6, 1:3) # > Error: `x` and `w` must be the same length
```

you can use `stopifnot` function to do multiple checks and get informative messages back

```{r}
wt_mean <- function(x, w, na.rm = FALSE) {
  stopifnot(is.logical(na.rm), length(na.rm) == 1)
  stopifnot(length(x) == length(w))

  if (na.rm) {
    miss <- is.na(x) | is.na(w)
    x <- x[!miss]
    w <- w[!miss]
  }
  sum(w * x) / sum(x)
}
```

`is.logical(na.rm)` is not TRUE

```{r, eval = FALSE}
wt_mean(1:6, 6:1, na.rm = "foo") #> Error: is.logical(na.rm) is not TRUE
```

Note that when using `stopifnot` you assert what should be true rather than checking for what might be wrong.

