## Overview

Looking at functions from e.g. `base` and `stats` packages that are loaded by default R.

## Allocation

as in allocation of empty objects / place holders

```{r "place holder allocation"}
vector("integer", 6)
vector("double", 7)
vector("character", 8)
vector("logical", 3)
```

## File Manipulation

Working with files. Checking existance, copy `cp`, removing `rm`, appending etc.

```{r "file-manipulation"}
library(magrittr)
tmpDir <- file.path("", "tmp", "r-file-manipulation")
dir.exists(tmpDir)
dir.create(tmpDir)
dir.exists(tmpDir)
list.files(tmpDir) # return empty character atomic vector

filePath1 <- tmpDir %>% file.path("A.md")
filePath2 <- tmpDir %>% file.path("B.md")
filePath3 <- tmpDir %>% file.path("C.md")
cat("file A\n", file = filePath1)
file.exists(filePath2)
cat("file B\n", file = filePath2)
file.exists(filePath2)
file.append(filePath1, filePath2)
file.info(filePath1)
# file.show(filePath1) # opens file, only run if in an interactive session
file.copy(filePath1, filePath3)
file.remove(filePath1, filePath2, filePath3)
file.remove(tmpDir) # works on EMPTY directories
```

## Extract

aka `[`. Use `drop` to keep the matrix.

```{r "base:extract"}
(m <- matrix(1:9, nrow = 3, dimnames = list(c("a", "b", "c"), LETTERS[1:3])))
m[1, ] # first row of m (atomic vector)
m[1, , drop = FALSE] # first row of m still as matrix
```

## replications

replicate elements of vectors and lists

```{r "base::rep"}
rep(1:4, 2)
rep(1:4, each = 2)       # not the same.
rep(1:4, c(2,2,2,2))     # same as second.
rep(1:4, c(2,1,2,1))
rep(1:4, each = 2, len = 4)    # first 4 only.
rep(1:4, each = 2, len = 10)   # 8 integers plus two recycled 1's.
rep(1:4, each = 2, times = 3)  # length 24, 3 complete replications

(x <- LETTERS[1:4] %>% setNames(letters[1:4]) %>% as.factor)
rep(x, times = 2)
rep(x, each = 2)
rep.int(x, times = 2)  # no names
rep(x, len = 10)
rep_len(x, length.out = 10) # no names
```

## iterate over data frame variables

use `base::seq_along` in a for loop to iterate over a data frame's variables / columns

```{r}
(scaledArrests <- USArrests) %>% head
for(i in seq_along(USArrests)){
  scaledArrests[[i]] <- scale(USArrests[[i]] %>% as.numeric) %>% round(3)
}
scaledArrests %>% head
```

be aware that the same thing (minus the row names) can be acheived much more nicely with `sapply`

```{r}
USArrests %>% dplyr::select(-state) %>% sapply(scale) %>% as.data.frame %>% head
```

## reduce function

`base::Reduce`

Common Higher-Order Functions in Functional Programming Languages (similar to Common Lisp's `(reduce)`)

Simple example

```{r}
add <- function(x) Reduce("+", x)
add(1:4)
```

Custom cumulative function. Approximation for pi. `right` argument determines if process starts from the right (or from the left).

```{r}
c(3, 7, 15, 1, 292) %>%
  Reduce(f = function(incr, acc) incr + 1 / acc, right = TRUE)
```

## pattern matching

`base::switch`

exact pattern matching in R

```{r}
centre <- function(x, type) {
  switch(type,
         mean = mean(x),
         median = median(x),
         trimmed = mean(x, trim = .1))
}
x <- rcauchy(10)
centre(x, "mean")
centre(x, "median")
centre(x, "trimmed")
```

## split df to list

e.g. split a `data.frame`, rows / observations to list. NOTE: second argument is a "group" e.g. `1:10`. If you in this case have a data frame with ten rows, each row will be in an entry in the returned list.

```{r "listify observations to iterate over rows"}
data.frame(a = 1:10
           ,b = 11:20
           ,c = 51:60) %>%
  split(seq(nrow(.)))
```

or by some other grouping

```{r}
split(airquality, airquality$Month)
```

## work inside closed data environment

`base::with` and `base::within`

```{r}
with(airquality, { cor(Temp, Ozone, use = "complete") })

aq <- within(airquality, {     # Notice that multiple vars can be changed
    lOzone <- log(Ozone)
    Month <- factor(month.abb[Month])
    cTemp <- round((Temp - 32) * 5/9, 1) # From Fahrenheit to Celsius
    S.cT <- Solar.R / cTemp  # using the newly created variable
    rm(Day, Temp)
})
head(aq)
```

## system information

invoke system calls / command line. check system info and library / package info.

```{r "OS calls"}
system("date")
system("echo $USER")

Sys.getenv()

Sys.info() %>% as.matrix

.libPaths()
```

useful for installing packages to a specific library / folder: `install.packages("magrittr", lib = .libPaths()[2])`

default lib parameter is `.libPaths()[1]` which tends to be user specific

`R_LIBS_USER` sets the personal library for a user.

`Sys.unsetenv("R_LIBS_USER")` can unset this so you use a global library path instead

R home

```{r}
R.home()
Sys.getenv("R_HOME")
```

## Citation

```{r message=FALSE}
require(magrittr)
```

```{r}
citation() %>% toBibtex # defaults to R-documentation
citation("magrittr") %>% toBibtex
```

## package version

check the version of an available package

```{r "packageVersion"}
packageVersion("magrittr")
```

## set reference level

`stats:relevel`

to set reference/base/stardard/default level for e.g. a logistic regression use `relevel` with the `ref` flag

```{r}
warpbreaks %>% str
warpbreaks$tension <- relevel(warpbreaks$tension, ref = "M")
warpbreaks %>% str
```

## scientific notation

print in scientific notation

```{r}
library(magrittr)
1234567890123
123456
123456 %>% format(scientific = TRUE)
```

you can also tinker with the `options(scipen=)` option

## print declared variable

use parentheses to return / print result during variable declaration

```{r}
(a <- 1)
a
```

## dicotomize

discretize continuous variable. (turn continuous variable discrete)

```{r}
Z <- stats::rnorm(10000)
table(cut(Z, breaks = -6:6))

breaks <- c(-3, 1, 5) # one more break than labels
labels <- c("low", "high")
U <- runif(10, min=-3, max=5)
cut(U,
    breaks = breaks,
    labels = labels,
    ordered_result = TRUE, # for ordered factors (not always well supported)
    include.lowest = TRUE)
```

## matching

`match` et. al match on the first character forward.

```{r}
1:10 %in% c(1,3,5,9)
charmatch("m",   c("mean", "median", "mode"))
charmatch("med", c("mean", "median", "mode"))
```

#### using grep

matching character strings/vectors using regexes or character strings. use `value` flag to return the actual matches and not the index.

```{r}
txt <- c("arm", "foot", "lefroo", "bafoobar")
grep("foo", txt)
grep("foo", txt, value = TRUE)
```

matching two exact expressions, either or

```{r}
library(magrittr)
txt <-
  c( "The", "licenses", "for", "most", "software", "are", "designed", "to", "take", "away", "your", "freedom", "to", "share"
    ,"and", "change", "it.", "", "By", "contrast,", "the", "GNU", "General", "Public", "License", "is", "intended", "to"
    ,"guarantee", "your", "freedom", "to", "share", "and", "change", "free", "software", "--", "to", "make", "sure"
    ,"the", "software", "is", "free", "for", "all", "its", "users")
```

returns index

```{r}
txt %>% grep(pattern = "ak")
txt %>% grep(pattern = "ak|an")
```

returns value

```{r}
txt %>% grep(pattern = "ak", value = TRUE)
txt %>% grep(pattern = "ak|an", value = TRUE)
```

returns logical (grep logical)

```{r}
txt %>% grepl(pattern = "ak")
txt %>% grepl(pattern = "ak|an")
```

matching on multiple character strings inside the same character string

```{r}
txt %>% grep(pattern = "s.*ar", value = TRUE)

txt %>% grep(pattern = "ra", value = TRUE)
txt %>% grep(pattern = "n.*ra", value = TRUE)
```

matching on end of string by using dollar sign `$` at the end of the pattern

```{r}
strs <-
  c("Product.Age", "Salesforce.Center", "Category", "Standard.Months",
    "FY15.10", "FY15.11", "FY15.12", "FY16.01", "FY16.02", "FY16.03",
    "FY16.04", "ofc-FY16.05", "ofc-FY16.06", "FY16.07", "ofc-FY16.08", "FY16.09",
    "FY16.10", "FY16.11", "FY16.12", "FY17.01", "FY17.02", "FY17.03",
    "FY17.04", "FY17.05", "FY17.06", "FY15.09.1",
    "FY15.10.1", "FY15.11.1", "FY15.12.1", "FY16.01.1", "FY16.02.1",
    "FY16.03.1", "FY16.04.1", "ofc-FY16.05.1", "ofc-FY16.06.1", "FY16.07.1",
    "FY16.08.1", "FY16.09.1", "FY16.10.1", "ofc-FY16.11.1", "ofc-FY16.12.1",
    "FY17.01.1", "FY17.02.1", "FY17.03.1", "FY17.04.1", "FY17.05.1",
    "FY17.06.1")
strs %>% grep(pattern = "(FY\\d{2}).(\\d{2})$", value = TRUE)
```

matching on start of string by using caret `^` at the start of the pattern

```{r}
strs %>% grep(pattern = "^(FY\\d{2}).(\\d{2})$", value = TRUE)
```

## padding

padding with zeros

```{r}
sprintf("%04d", 20)
```

or less stable

```{r}
sprintf("%02s", c("10", "8", "12"))
```

case with hours and minutes, similar to a clock

```{r "sprintf clock"}
sprintf("%02s:%02s", 0:5, 15)
sprintf("%02s:%02s", 0:5, c(5,25,55)) # not expected
sprintf("%02s:%02s", rep(0:5, each = 3), c(5,25,55)) # expected

minutesAsClock <-
  function(minutes)
    sprintf( "%02s:%02s"
            ,minutes %>% magrittr::divide_by_int(60)
            ,minutes %>% magrittr::mod(60)) # modulo `%%`
(0:12*60+40) %>% minutesAsClock
```

Windows

```{r "sprintf on windows"}

sprintf("%02s", c("10", "8", "12")) # doesn't return what you'd expect on windows >.<
# sprintf("%02d", c("10", "8", "12")) # however this does! using `d`
# not sure it works as intended on Unix systems though

```

## substr

can extract characters based on position

```{r}
c('201611', '201304') %>% substr(5,6)
c('201611', '201304') %>% substr(1,4)
```

Can be working similar to T-SQL's `LEFT`-function. No need to use regex

Can also be thought of as deleting parts of strings / characters.

## numerical characteristics of the machine

`.Machine` has a lot of numerical characteristics

e.g. the rounding error

```{r}
.Machine$double.eps
.Machine
```

## setting names for arrays

use `stats::setNames`

```{r}
setNames( 1:3, c("foo", "bar", "baz") )
setNames( 1:3, nm = c("foo", "bar", "baz") )
setNames( nm = c("foo", "bar", "baz") ) # Special case
```

or `base::structure`; NOTE: you need to name any arguments, unlike `setNames`.

```{r}
structure( 1:3, nm = c("foo", "bar", "baz") ) # don't use `nm`
structure( 1:3, names = c("foo", "bar", "baz") )
# structure( 1:3, c("foo", "bar", "baz") ) # Error in structure(1:3, c("foo", "bar", "baz")) : attributes must be named
```

## scale

for normalizing / standardizing variables in `data.frame`

`scale` takes three arguments, the data, center and scale.

- `center` what you subtract from the values (unless set to `FALSE`). Default is the mean of the column in question.
- `scale` is the demoninator (unless set to `FALSE`). Default is the standard deviation.

```{r "scale function"}

airquality %>% summary

airquality %>% scale %>% summary # NOTE: the means and symmetrics between min/max

```

function to scale columns to `[0,1]` unit range

TODO: add scale function to go between `[-1,1]`

```{r}

unitScale <- function(dta)
  scale(dta
        ,center = dta %>% purrr::map_dbl(min, na.rm = TRUE)
        ,scale = dta %>% purrr::map(range, na.rm = TRUE) %>% purrr::map_dbl(diff)
        )
airquality %>% unitScale %>% summary

```

```{r regexes, child="regexes.Rmd", cache=FALSE}
```

