
---
title: "Ruby cheatsheet"
author: Emil Rehnberg
bibliography: refs.bib
csl: shiki.csl
output:
  pdf_document:
    highlight: zenburn
  html_document:
    toc_float: TRUE
    css: styles.css
---

```{r set-options, echo=FALSE, cache=FALSE}
options(width = 200)
```

```{r echo=FALSE, eval=FALSE}
library(shiny)
rmdFilePath <- "ruby.rmd"
rmarkdown::render(rmdFilePath, output_format="html_document") # "all_document"
```

```{r echo=FALSE, message=FALSE}
set.seed(308)
library(magrittr)
```

## 目的

Cheatsheet for Ruby. It's a place to dump typical annotated Ruby-code.

## Tricks { .tabset .tabset-fade .tabset-pills }

### Overview

Tricks in vanilla ruby.

### Logging to file

```rb
#!/usr/bin/env ruby

if __FILE__ == $0
  $stdout.reopen("out.txt", "w")
  $stderr.reopen("err.txt", "w")

  puts "runing R sesh"
  puts %x{Rscript rr.R}
  "kenny florian finish fights"
end
```

perhaps in combination with `rr.R`

```R
#!/bin/usr/env Rscript

sink(stdout(), type="message")
message("message: testing 12")
warning("achtung: deutchen furher")
error("E$ failure") # doesn't work, using `stop` if you want to send error messages and stop the execution.
```

[ruby source](https://stackoverflow.com/questions/3018595/how-do-i-redirect-stderr-and-stdout-to-file-for-a-ruby-script)
[r source](http://mazamascience.com/WorkingWithData/?p=888)

### class accessor

this is possible but not ideal (Avdi)

```rb
class MyLib
  class << self
    attr_accessor :logger
  end
end
MyLib.logger = Logger.new($stderr)
```

instead write the accessors yourself!

```rb
class MyLib
  def self.logger
    @logger ||= Logger.new($stderr)
  end

  def self.logger=(new_logger)
    @logger = new_logger
  end
end
```

### all combinations for a vector

```rb
a = [1, 2, 3]
a
  .each_index
  .map(&a.method(:rotate))
```

## rspec { .tabset .tabset-fade .tabset-pills }

### Overview

section for rspec stuff.

### tags

in `greeter_spec`

```rb
  it 'says "Hello"', :wip do
    raise unless greeter.msg == 'Hello'
  done
```

only exec specs wip tags

```sh
rspec --tag wip greeter_spec.rb
```

use `~wip` to run specs w/o the `wip` tag

## sucker-punch

concurrent execution in ruby. based on `concurrent-ruby`

```rb
#!/usr/bin/env ruby

require 'logger'
require 'sucker_punch'

class Shirshasana
  def perform
    puts "stay 24 breaths. should be 12 mins in total"
    sleep(1)
  end
end

class Padmasana
  def perform
    puts "stay 10 breaths. count the breaths using your mala."
    sleep(1)
  end
end

class AsanaPractioner
  ASANA_LOOKUP ={
    shirshasana: Shirshasana.new,
    padmasana: Padmasana.new
  }

  def initialize(asana_name)
    @asana = ASANA_LOOKUP[asana_name]
  end

  def perform
    @asana.perform
  end
end

class Practice
  include SuckerPunch::Job
  workers 2

  def perform(asana_name, log)
    log.debug(AsanaPractioner.new(asana_name).perform)
  end
end

if __FILE__ == $0

  puts "performing shirshasana"
  shirshasana_log = Logger.new("/tmp/shirshasana.log", 'monthly')
  Practice.perform_async(:shirshasana, shirshasana_log)
  Practice.perform_async(:shirshasana, shirshasana_log)
  Practice.perform_async(:shirshasana, shirshasana_log)
  Practice.perform_async(:shirshasana, shirshasana_log)
  Practice.perform_async(:shirshasana, shirshasana_log)

  puts "performing padmasana"
  padmasana_log = Logger.new("/tmp/padmasana.log", 'monthly')
  Practice.perform_async(:padmasana, padmasana_log)
  Practice.perform_async(:padmasana, padmasana_log)
  Practice.perform_async(:padmasana, padmasana_log)
  Practice.perform_async(:padmasana, padmasana_log)

  puts "finished"
end
```


