
---
title: "SQL cheatsheet"
author: Emil Rehnberg
bibliography: refs.bib
csl: shiki.csl
output:
  pdf_document:
    highlight: zenburn
  html_document:
    toc_float: TRUE
    css: styles.css
---

```{r set-options, echo=FALSE, cache=FALSE}
options(width = 200)
```

```{r echo=FALSE, eval=FALSE}
library(rmarkdown); library(shiny)
rmdFilePath <- "sql.rmd"
rmarkdown::render(rmdFilePath, output_format="html_document") # "all_document"
```

```{r echo=FALSE, message=FALSE}
set.seed(308)
library(magrittr)
```

## 目的

Cheatsheet for SQL. It's a place to dump typical annotated SQL-code.

### General / Tricks

Sections for ANSI, Transact-SQL, Postgresql, MySQL

### ANSI { .tabset .tabset-fade .tabset-pills }

#### Overview

Standard SQL and "common" SQL (e.g. window functions)

#### Rollbacks

- use `BEGIN TRAN(SACTION)` to open a transaction
- `BEGIN TRAN T1` names the transaction `T1`
    - use-case: applying nested/multiple transactions
- using `WITH MARK [descr]` with named transaction makes the (named) log restorable

```sql
BEGIN TRAN T1;
  UPDATE table1 ...;
  BEGIN TRAN M2 WITH MARK;
    UPDATE table2 ...;
  COMMIT TRAN M2;
  UPDATE table3 ...;
COMMIT TRAN T1;
```

#### Find first missing key/number

Main idea: use <code>NOT EXISTS</code> to <i>catch</i> when <i>key + 1 does not exist</i> ( and `NOT EXISTS` catches this)

```sql
SELECT
  CASE
  WHEN NOT EXISTS(
    SELECT *
    FROM dbo.T1
    WHERE key = 1)
  THEN 1
  ELSE (...subquery for min NA...)
  END AS ...
  ...
```

Returns 1 if 1 doesn't exist, otherwise, it returns the subquery result. The subquery returns the minimum missing value &gt;1

```sql
SELECT  MIN(A.key) + 1 as missing
FROM    dbo.T1 AS A
WHERE NOT EXISTS(
  SELECT *
  FROM dbo.T1 AS B
  WHERE B.key = A.key + 1)
```

The `NOT EXISTS` predicate returns `TRUE` only for values in `T1` that are right before a gap.

A nice solution using `OUTER JOIN`.

- Identify `NULL`s in outer rows for smallest missing key (keys > 1)

```sql
SELECT MIN(A.keycol) + 1
FROM dbo.T1 AS A
  LEFT OUTER JOIN dbo.T1 AS B
    ON B.keycol = A.keycol + 1
WHERE B.keycol IS NULL;
```

To put your knowledge of the `EXISTS` predicate into action, try to solve the following problem. Notice that keycol must be positive

```sql
USE tempdb;
GO
  IF OBJECT_ID('dbo.T1') IS NOT NULL
    DROP TABLE dbo.T1;
GO

CREATE TABLE dbo.T1
(
  keycol  INT         ...,
  datacol VARCHAR(10) ...
);
INSERT INTO dbo.T1(keycol, datacol)
     VALUES
  (3, 'a'),
  (4, 'b'),
  (6, 'c'),
  (7, 'd');
```

Write a query that returns the lowest missing key (starting at 1). E.g. the table is currently populated with the keys 3, 4, 6, and 7. Query should return the value 1. If you insert two more rows, with the keys 1 and 2, your query should return 5.

Alternative solution (perhaps harder to read)

```sql
SELECT COALESCE(MIN(A.keycol) + 1, 1)
FROM dbo.T1 AS A
WHERE
  NOT EXISTS(
    SELECT * FROM dbo.T1 AS B
    WHERE B.keycol= A.keycol + 1)
  AND EXISTS(
    SELECT * FROM dbo.T1
    WHERE keycol = 1);
```

#### Removing duplicates with DELETE

use `ROW_NUMBER` over unique traits and `DELETE` where row number is over 1

```sql
WITH Dups AS (
  SELECT orderid, custid, empid, orderdate,
    ROW_NUMBER()
      OVER(PARTITION BY orderid
        ORDER BY (SELECT 0)) AS rn
  FROM dbo.OrdersDups
)
DELETE FROM Dups
WHERE rn > 1;
```

calculate `ROW_NUMBER` and `RANK` over unique traits and delete where they differ

```sql
WITH Dups AS (
  SELECT orderid, custid, empid, orderdate,
    ROW_NUMBER()
      OVER(PARTITION BY orderid
        ORDER BY (SELECT 0)) AS rn,
    RANK()
      OVER(PARTITION BY orderid
        ORDER BY (SELECT 0)) AS rnk
  FROM dbo.OrdersDups
)
DELETE FROM Dups
WHERE rn <> rnk;
```

#### last order for each customer

Extract last order for each customer
standard SQL

```sql
SELECT id, customer_id, order_date, shipping_country
  FROM orders o1
  JOIN (
    SELECT customer_id     AS cid,
           MAX(order_date) AS maxdate
      FROM orders
  GROUP BY customer_id
       ) o2
    ON o1.customer_id = o2.cid
   AND o1.order_date = o2.{{maxdate}};
```

#### Relational division

return orders that contain a basket with products 2, 3, and 4.

```sql
SELECT orderid
FROM (SELECT
        orderid,
        COUNT(CASE WHEN productid = 2 THEN productid END) AS P2,
        COUNT(CASE WHEN productid = 3 THEN productid END) AS P3,
        COUNT(CASE WHEN productid = 4 THEN productid END) AS P4
      FROM dbo.OrderDetails
      GROUP BY orderid) AS P
WHERE P2 = 1 AND P3 = 1 AND P4 = 1;
```

derived table returns `A` and `B` rows from

```
order_id P2   P3   P4
-------- ---- ---- --
A        1    1    1
B        1    1    1
C        NULL 1    1
D        NULL NULL 1
```

#### Logical and bit

for logicals expression in BIT to access tools yielding True/False

```sql
SELECT
  CAST(CASE WHEN [lf].[InvoiceDimDateId] > 0 THEN 1
            ELSE 0
       END
       AS BIT)
FROM dbo.Order
```

### Postgresql { .tabset .tabset-fade .tabset-pills }

#### Overview

open source dialect, boosting standard ANSI sql.

#### last order for each customer

Extract last order for each customer
standard SQL

```sql
  SELECT DISTINCT ON(customer_id)
         *
    FROM orders
ORDER BY customer_id::remember last order,
         order_date DESC;
```

```sql
 id | customer_id | order_date | shipping_country
----+-------------+------------+------------------
  3 |           2 | 2015-12-12 | NO
  2 |           4 | 2015-10-05 | JA
(2 rows)
```

### T-SQL { .tabset .tabset-fade .tabset-pills }

#### Overview

MS sql dialect transact-sql for SQL server

#### Date to integer convertion

to convert a datetime or date column/value to a dateId in integer use the `CONVERT` function

```sql
CONVERT(VARCHAR(8), StartTime, 112)
```

#### Triggers

triggers work like callback::programming terms. The following prints 'INSERT detected' after an `INSERT`

```sql
CREATE TRIGGER trgCustsINSERT
  ON Custs
  AFTER INSERT
AS
  PRINT 'INSERT detected.';
GO
```

#### Adding DATETIME and TIME

```sql
DECLARE @d DATETIME = '2013-02-18T18:34:40.330',
        @t TIME(7)  = '00:11:00.0000000';

SELECT DATEADD(
  SECOND,
  DATEDIFF(SECOND, 0, @t),
  @d);
```

```sql
...
-----------------------
2013-02-18 18:45:40.330
```

#### partitioned views

two orthogonal classifications

- physical placement of the component tables
    - local: all tables constituting a partitioned view are located in a single instance of SQL Server
    - distributed: tables are located across two or more instances
- updatability
    - updatable: has a single column constraint on each participating table to determine which table a row belongs to
    - not updatable: -

#### Comparing Partitioned Views / Tables

partitioned table
parts must reside in the same DB
must have same constraints and primary keys
partitioned view
can be split up over servers and machines
can be slower and quicker than table defending on the systems
optimazation is typically more complex
more objects

#### create empty table

```sql
SELECT TOP (0) column_list
INTO table
FROM other_table
```

#### emptying a table

keep table columns, reset ID counting, no logging.

```sql
TRUNCATE TABLE tableName;
```

#### archive deleted records

archive deleted data with `DELETE OUTPUT (INTO)`

```sql
WHILE 1 = 1
BEGIN
  DELETE TOP(5000) FROM orders
    OUTPUT deleted.orderid, deleted.date
      INTO archive(orderid, date)
  WHERE date < '20060101';
  IF @@rowcount < 5000 BREAK;
END
```

#### delete lots of rows in a table

if you want delete lots of rows in a table

- TRUNCATE TABLE not doing it
- want logging
- want to keep a few rows
- technique is useful when you want to archive deletions w DELETE OUTPUT

```sql
DELETE in batches
WHILE 1 = 1
BEGIN
  DELETE TOP (5000)
  FROM orders
  WHERE orderdate < '20060101';
  IF @@rowcount <= 5000 BREAK;
END
```

#### Apply clause

The `APPLY` operator involves one or two steps:

1. Apply Right Table Expression to Left Table Rows
2. Add Outer Rows

`CROSS APPLY` uses 1., `OUTER APPLY` uses both.

I think of `APPLY` as a way to loop over a set of rows (something which might be nice when you want to match on a set element-by-element)

query returning the two (highest order IDs) orders for each
customer:

```sql
SELECT C.customerid, C.city, A.orderid
  FROM dbo.Customers AS C
  CROSS APPLY
    (SELECT TOP (2) O.orderid, O.customerid
       FROM dbo.Orders AS O
      WHERE O.customerid = C.customerid
   ORDER BY orderid DESC) AS A;
```

This query generates the following output:

```sql
custid city   orderid
------ ------ -------
FRNDO  Madrid 2
FRNDO  Madrid 1
KRLOS  Madrid 5
KRLOS  Madrid 4
MRPHS  Zion   6
```

Notice that FISSA is missing from the output because the table expression A returned an empty set for it. If you also want to return customers who placed no orders, use `OUTER APPLY` as follows:

```sql
SELECT C.customerid, C.city, A.orderid
FROM dbo.Customers AS C
  OUTER APPLY
    (SELECT TOP (2) O.orderid, O.customerid
     FROM dbo.Orders AS O
     WHERE O.customerid = C.customerid
     ORDER BY orderid DESC) AS A;
```

This query generates the following output:

```sql
custid city   orderid
------ ------ -------
FISSA  Madrid NULL
FRNDO  Madrid 2
FRNDO  Madrid 1
KRLOS  Madrid 5
KRLOS  Madrid 4
MRPHS  Zion   6
```

#### Grouping ID

- helps with identifying grouped rows
    - instead of matching on `NULL`s
- accepts a list of attributes and constructs an integer bitmap for each attribute
    - bit is 0 grouping set member attributes, 1 otherwise
- the rightmost bit represents the rightmost input attribute
- `GROUPING_ID(a, b, c, d)` returns
    - 0 (0×8 + 0×4 + 0×2 + 0×1) for grouping set (a, b, c, d)
    - 1 (0×8 + 0×4 + 0×2 + 1×1) for grouping set (a, b, c)
    - 2 (0×8 + 0×4 + 1×2 + 0×1) for grouping set (a, b, d)
    - 3 (0×8 + 0×4 + 1×2 + 1×1) for grouping set (a, b)
    - ...

```sql
SELECT
  GROUPING_ID(
    custid,
    empid,
    YEAR(date),
    MONTH(date),
    DAY(date) ) AS grp_id,
  c_id, e_id,
  YEAR(date) AS year,
  MONTH(date) AS mnth,
  DAY(date) AS day,
  SUM(qty) AS qty
FROM dbo.Orders
GROUP BY
  CUBE(c_id, e_id),
  ROLLUP(YEAR(date), MONTH(date), DAY(date))
```

`grp_id` value 25 is grouping set `(year, mnth)`. I.e. the second and third bits (value 2,4).

```sql
grp_id c_id e_id year mnth day  qty
------ ---- ---- ---- ---- ---- ---
0      C    3    2006 4    18   22
16     NULL 3    2006 4    18   22
0      A    3    2006 8    2    10
24     NULL NULL 2006 4    18   22
25     NULL NULL 2006 4    NULL 22
16     NULL 3    2006 8    2    10
24     NULL NULL 2006 8    2    10
25     NULL NULL 2006 8    NULL 10
0      D    3    2006 9    7    30
16     NULL 3    2006 9    7    30
...
```

#### INSERT EXEC stored procedure

```sql
CREATE PROC dbo.GetRows
  @n AS INT = 10
AS
SELECT ...
...

CREATE PROC ...
  @to_rownum AS INT ...
AS
...
INSERT
INTO #CachedPages
    EXEC dbo.GetRows
      @n = @to_rownum::set proc argument @n to @to_rownum;
...
```

#### PIVOT clause

- operator allowing you to rotate, or pivot, data between columns and rows, performing aggregations along the way.
- requires alias
    - will otherwise return Incorrect syntax near ';'.

```sql
SELECT *
  FROM (
  SELECT
     empid,
     YEAR(orderdate) AS year,
     val
    FROM Sales.OrderValues)
 PIVOT(SUM(val)
  FOR year
   IN([2006],[2007],[2008]) AS P);
```

query generates the following

```
empid 2006     2007      2008
----- -------- --------- --------
3     18223.96 108026.17 76562.75
6     16642.61 43126.38  14144.16
9     9894.52  26310.39  41103.17
7     15232.16 60471.19  48864.89
1     35764.52 93148.11  63195.02
4     49945.12 128809.81 54135.94
2     21757.06 70444.14  74336.56
5     18383.92 30716.48  19691.90
8     22240.12 56032.63  48589.54
```

e.g. query for the total value of orders handled by each employee for each order year. output to have a row for each employee, a column for each order year, and the total value in the intersection of each employee and year.

#### Relational division

return orders that contain a basket with products 2, 3, and 4.
using PIVOT:

```sql
SELECT orderid
FROM (
  SELECT orderid, productid
  FROM dbo.OrderDetails) AS D
  PIVOT(COUNT(productid)
    FOR productid
    IN([2],[3],[4])) AS P
WHERE [2] = 1 AND [3] = 1 AND [4] = 1;
```

derived table returns `A` and `B` rows from

```
order_id P2   P3   P4
-------- ---- ---- --
A        1    1    1
B        1    1    1
C        NULL 1    1
D        NULL NULL 1
```

### SQLite { .tabset .tabset-fade .tabset-pills }

#### Overview

Simple embedded sql dialect. Usable in just about any embedded software!!

#### run code for db

from command line

```sql
sqlite3 ex2.db < ex2.sql
```

#### print schema from db

from command line

```sql
sqlite3 ex2.db .schema
```

```sql
CREATE TABLE person (
  id INTEGER PRIMARY KEY,
  first_name TEXT,
  last_name TEXT,
  age INTEGER
);
CREATE TABLE pet (
  id INTEGER PRIMARY KEY,
  name TEXT,
  breed TEXT,
  age INTEGER,
  dead INTEGER
);
CREATE TABLE person_pet (
  person_id INTEGER,
  pet_id INTEGER
);
```


