
---
title: "stats cheatsheet"
author: Emil Rehnberg
bibliography: refs.bib
csl: shiki.csl
output:
  pdf_document:
    highlight: zenburn
  html_document:
    toc_float: TRUE
    css: styles.css
---

```{r set-options, echo=FALSE, cache=FALSE}
options(width = 200)
```

```{r echo=FALSE, eval=FALSE}
require(rmarkdown); require(shiny)
rmdFilePath <- "stats.rmd"
rmarkdown::render(rmdFilePath, output_format="html_document") # "all_document"
```

```{r echo=FALSE, message=FALSE}
set.seed(308)
require(colorout)
require(magrittr)
```

```{r echo=FALSE}
# abbreviation tags
dlmAbbrTag <- tags$abbr("DLM", title="Dynamic Linear Model")
```

## 目的

cheatsheet for statistics. it's a place to dump thoughtworthy stuff that I might want to look up for reminders.

## Methodologies {.tabset .tabset-fade .tabset-pills}

### DLMs

#### Definition [@west1999multiDLM]

`r dlmAbbrTag`: For each $t$, the model is defined by
<dl>
  <dt>Observation equation:</dt>
  <dd>$Y_t = \mathbf{F}'_t \boldsymbol\theta_t + \nu_t, \; \nu_t \sim \mathsf{N}[0, V_t]$</dd>
  <dt>System equation:</dt>
  <dd>$\boldsymbol\theta_t = \boldsymbol\theta_{t-1} + \boldsymbol\omega_t, \;
  \boldsymbol\omega_t \sim \mathsf{N}[\mathbf{0}, \mathbf{W}_t]$</dd>
</dl>

- **regression vector**: $\mathbf{F}_t = (X_{t1} ,\ldots, X_{tn})'$ (predictors)
- **regression parameter vector**: $\boldsymbol\theta_t$ ($n \times 1$) (coefficients)
- **evolution variance matrix**: $\mathbf{W}_t$ (for $\boldsymbol\theta_t$)
    - the evolution error term $\boldsymbol\omega_t$ describes the changes in the elements of the parameter vector between times $t-1$ and $t$.
    - the zero mean vector reflects that $\boldsymbol\theta_t$ is expected to be constant over the interval
    - the variance matrix $\mathbf{W}_t$ governs the extent of the movements in $\boldsymbol\theta_t$ and hence the extent of the time period over which the assumption of local contancy is reasonable

illustration of when $n = 1$ yields a straight line regression
$\mathbf{F}_t = (1, X_t)'$ and $\boldsymbol\theta_t = (\alpha_t, \beta_t)'$.
Then
$$
\begin{align*}
  Y_t & = \alpha_t + \beta_t X_t + \nu_t, & \nu_t & \sim \mathsf{N}(0, V_t) \\
  \alpha_t & = \alpha_{t-1} + \omega_{t1}  \\
  \beta_t & = \beta_{t-1} + \omega_{t2}
\end{align*}
$$

where $\boldsymbol\omega_t = (\omega_{t1}, \omega_{t2})' \sim \mathsf{N}[0, \mathbf{W}_t]$

#### General normal definition

Set of params definines DLM $(\mathbf{F}, \mathbf{G}, \mathbf{V}, \mathbf{W})$ [@west1999generalDLMdef]

where

- $r$ - number of variables to estimate e.g. timeslots
- $n$ - number of parameters to fit and use e.g. seasonality parts

at each time $t$ ()

$$
\begin{align*}
  \mathbf{F}_t & \mbox{ is a known } (n \times r) \mbox{ matrix}\\
  \mathbf{G}_t & \mbox{ is a known } (n \times n) \mbox{ matrix}\\
  \mathbf{V}_t & \mbox{ is a known } (r \times r) \mbox{ matrix}\\
  \mathbf{W}_t & \mbox{ is a known } (n \times n) \mbox{ matrix}
\end{align*}
$$

this quadruple defines the `r dlmAbbrTag` relating $\mathbf{Y}_t$ to the $(n \times 1)$ parameter $\boldsymbol\theta_t$ at time $t$ and
$$
\begin{align*}
  (\mathbf{Y}_t | \boldsymbol\theta_t) & \sim N[\mathbf{F}_t' \boldsymbol\theta_t, \mathbf{V}_t] \\
  (\boldsymbol\theta_t | \boldsymbol\theta_{t-1}) & \sim N[\mathbf{G}_t' \boldsymbol\theta_{t-1}, \mathbf{W}_t]
\end{align*}
$$

or

$$
\begin{align*}
  \mathbf{Y}_t & = \mathbf{F}_t'\boldsymbol\theta_t + \boldsymbol\nu_t, &
  \boldsymbol\nu_t \sim & N[\mathbf{0}, \mathbf{V}_t] & \mbox{ (aka observation equation)} \\
  \boldsymbol\theta_t & = \mathbf{G}_t'\boldsymbol\theta_{t-1} + \boldsymbol\omega_t, &
  \boldsymbol\omega_t \sim & N[\mathbf{0}, \mathbf{W}_t] & \mbox{ (aka evolution, state or system equation)}
\end{align*}
$$

- $\mathbf{F}_t$ is the design matrix of known values of independent variables
- $\boldsymbol\theta_t$ is the state, or system, vector
- $\boldsymbol\mu_t = \mathbf{F}'_t \boldsymbol\theta_t$ is the mean response or level
- $\boldsymbol\nu_t$ is the observation error
- $\mathbf{G}_t$ is the evolution, system, transfer or state matrix
- $\boldsymbol\omega_t$ is the system, or evolution, error with evolution variance $\mathbf{W}_t$

Parameter setting [@west1999obsVar]

- regression vectors $\mathbf{F}_t$ and evolution matrix $\mathbf{G}_t$ area defined by the modeller via model design principles (ch 5/6?)
- $\mathbf{W}_t$ is usually definied with the discount principle (ch 6) (discount factors $\delta$?)
- $V_t$ is often unknown and relatively large to system variance $\mathbf{W}_t$

one-step forecast [@west1999generalDLMdef]

1. Posterior for $\boldsymbol\theta_{t-1}$ (for some $\mathbf{m}_{t-1}$, and $\mathbf{C}_{t-1}$):
   $$ (\boldsymbol\theta_{t-1} | D_{t-1}) \sim \mathsf{N}[\mathbf{m}_{m-1}, \mathbf{C}_{t-1}] $$
2. Prior for $\boldsymbol\theta_t$:
   $$ (\boldsymbol\theta_{t} | D_{t-1}) \sim \mathsf{N}[\mathbf{a}_t, \mathbf{R}_{t}] \\
   \mbox{where } \mathbf{a}_t = \mathbf{G}_t\mathbf{m}_{t-1} \mbox{ and } \mathbf{R}_t = \mathbf{G}_tC_{t-1}\mathbf{G}'_t + \mathbf{W}_t$$
3. 1-step forecast:
   $$ (Y_t | D_{t-1}) \sim \mathsf{N}[f_t, Q_t] \\
   \mbox{where } f_t = \mathbf{F}'_t\mathbf{a}_t \mbox{ and } Q_t = \mathbf{F}'_tR_t\mathbf{F}_t + V_t$$
4. Posterior for $\theta_t$:
   $$ (\boldsymbol\theta_{t} | D_{t}) \sim \mathsf{N}[\mathbf{m}_t, \mathbf{C}_t] \\
   \mbox{with } \mathbf{m}_t = \mathbf{a}_t + \mathbf{A}_te_t \mbox{ and } \mathbf{C}_t = \mathbf{R}_t - \mathbf{A}_t Q_t\mathbf{A}'_t \\
   \mbox{where } \mathbf{A}_t = \mathbf{R}_t\mathbf{F}_t Q_t^{-1} \mbox{ and } e_t=Y_t - f_t$$

#### One-step forecast [@west1999oneStepDlm]

1. Posterior for $\theta_t$:
   $$ (\theta_{t-1} | D_{t-1}) \sim \mathsf{N}[m_{m-1}, C_{t-1}] \\
   \mbox{ for some mean } m_{t-1} \mbox{ and variance } C_{t-1}$$
2. Prior for $\theta_t$:
   $$ (\theta_{t} | D_{t-1}) \sim \mathsf{N}[m_{m-1}, R_{t}] \\
   \mbox{where } R_t = C_{t-1} + W_t$$
3. 1-step forecast:
   $$ (Y_t | D_{t-1}) \sim \mathsf{N}[f_t, Q_t] \\
   \mbox{where } f_t = F_tm_{t-1} \mbox{ and } Q_t = F^2_tR_t + V_t$$
4. Posterior for $\theta_t$:
   $$ (\theta_{t} | D_{t}) \sim \mathsf{N}[m_{m}, C_{t}] \\
   \mbox{with } m_t = m_{t-1} + A_te_t \mbox{ and } C_t = R_tV_t / Q_t \\
   \mbox{where } A_t = R_tF_t / Q_t \mbox{ and } e_t=Y_t - f_t$$

Comments

- The coefficient $A_t = R_tF_t / Q_t$ scales the correction term according to the relative precisions of the prior and likelihood, as measured by $R_t / Q_t$, and by the regressor value $F_t$. The correction always take the sign of $F_t$ and may be unbounded.
- Posterior precision
  $$ C^{-1}_t = Q_t(R_tV_t)^{-1} = R^{-1}_t + F^2_tV_t^{-1} $$
  is hence larger than $R^{-1}_t$ (unless $F_t = 0$ - $Y_t$ provides no information on $\theta_t$).
  Thus, the posterior for $\theta_t$ is never more diffuse than it's prior.
  If $F_t = 0$ for a sequence of observations then the $C_t$ grows by the addition of furhter $W_t$ terms, reflecting an increasingly diffuse posterior.

#### Discount factor

$\delta$ is often used as notation for the discount factor.

- $\delta\in[0,1]$ represents the difference in importance between future rewards and present rewards.
- inverse multiplied by $C$ to make $R$. (GL set-up)

### Overview

data mining, prediction stuff


TODO

- look up Kalman filtering? seems very standard
