
---
title: "stats cheatsheet"
author: Emil Rehnberg
bibliography: refs.bib
csl: shiki.csl
output:
  pdf_document:
    highlight: zenburn
  html_document:
    toc_float: TRUE
    css: styles.css
---

```{r set-options, echo=FALSE, cache=FALSE}
options(width = 200)
```

```{r echo=FALSE, eval=FALSE}
require(rmarkdown); require(shiny)
rmdFilePath <- "stats.rmd"
rmarkdown::render(rmdFilePath, output_format="html_document") # "all_document"
```

```{r echo=FALSE, message=FALSE}
set.seed(308)
require(colorout)
require(magrittr)
```

```{r echo=FALSE}
# abbreviation tags
dlmAbbrTag <- tags$abbr("DLM", title="Dynamic Linear Model")
```

## 目的

cheatsheet for statistics. it's a place to dump thoughtworthy stuff that I might want to look up for reminders.

## Methodologies {.tabset .tabset-fade .tabset-pills}

```{r eval=FALSE}
### Overview

data mining, prediction stuff
```

### DLMs

Definition of `r dlmAbbrTag` [@west1999multiDLM]

For each $t$, the model is defined by
<dl>
  <dt>Observation equation:</dt>
  <dd>$Y_t = \mathbf{F}'_t \boldsymbol\theta_t + \nu_t, \; \nu_t \sim \mathsf{N}[0, V_t]$</dd>
  <dt>System equation:</dt>
  <dd>$\boldsymbol\theta_t = \boldsymbol\theta_{t-1} + \boldsymbol\omega_t, \;
  \boldsymbol\omega_t \sim \mathsf{N}[\mathbf{0}, \mathbf{W}_t]$</dd>
</dl>

- **regression vector**: $\mathbf{F}_t = (X_{t1} ,\ldots, X_{tn})'$ (predictors)
- **regression parameter vector**: $\boldsymbol\theta_t$ ($n \times 1$) (coefficients)
- **evolution variance matrix**: $\mathbf{W}_t$ (for $\boldsymbol\theta_t$)
    - the evolution error term $\boldsymbol\omega_t$ describes the changes in the elements of the parameter vector between times $t-1$ and $t$.
    - the zero mean vector reflects that $\boldsymbol\theta_t$ is expected to be constant over the interval
    - the variance matrix $\mathbf{W}_t$ governs the extent of the movements in $\boldsymbol\theta_t$ and hence the extent of the time period over which the assumption of local contancy is reasonable

illustration of when $n = 1$ yields a straight line regression
$\mathbf{F}_t = (1, X_t)'$ and $\boldsymbol\theta_t = (\alpha_t, \beta_t)'$.
Then
$$
\begin{align*}
  Y_t & = \alpha_t + \beta_t X_t + \nu_t, & \nu_t & \sim \mathsf{N}(0, V_t) \\
  \alpha_t & = \alpha_{t-1} + \omega_{t1}  \\
  \beta_t & = \beta_{t-1} + \omega_{t2}
\end{align*}
$$

where $\boldsymbol\omega_t = (\omega_{t1}, \omega_{t2})' \sim \mathsf{N}[0, \mathbf{W}_t]$

