---
title: "unix cheatsheet"
author: Emil Rehnberg
bibliography: refs.bib
csl: shiki.csl
output:
  pdf_document:
    highlight: zenburn
  html_document:
    toc_float: TRUE
    css: styles.css
---

```{r set-options, echo=FALSE, cache=FALSE}
options(width = 200)
```

```{r echo=FALSE, eval=FALSE}
require(rmarkdown); require(shiny)
rmdFilePath <- "unix.rmd"
allexamples <- FALSE
rmarkdown::render(rmdFilePath, output_format="html_document") # "all_document"
```

## 目的

cheatsheet for unix.

## scheduled jobs {.tabset .tabset-fade .tabset-pills}

### Overview

section for scheduled jobs, aka cron jobs

### cron format

5 / 6 digits indicating the timing of the job

TODO: explain this further

### edit jobs

```sh
crontab -e
```

## users and groups {.tabset .tabset-fade .tabset-pills}

### Overview

[how-to-add-users](https://help.ubuntu.com/community/AddUsersHowto)

### add user

```sh
sudo adduser <username>
```

### add user to group

```sh
sudo addgroup <groupname>
```

### add user to existing group

```sh
sudo adduser <username> <groupname>
```

### show group members

```sh
getent group <groupname>
```

## workflows {.tabset .tabset-fade .tabset-pills}

### public keys setup for ssh

use private / public rsa keys to make ssh-logins more convenient :D

```sh
ssh-keygen
ssh-add ~/.ssh/id_rsa_password
ssh-copy-id -i ~/.ssh/id_rsa_password wms@de-2204.d.ipeer.se
```

### file compression

in order to compress a folder into a zip file

```sh
zip pack.zip folder-with-files    # just adds the folder!
zip -r pack.zip folder-with-files # add the r(ecursive) flag to zip the files inside the folder
```

