### Overview

if new to factors then check out the [section](http://r4ds.had.co.nz/factors.html) in R for data science.

### generate factor levels

```{r}
gl(2, 3, labels = c("Control", "Treatment"))
```

### drop factor levels

Use the `droplevels` function. Returns the full argument object with pruned factor levels.

```{r}
df <-
  data.frame(letters=letters[1:5],
             numbers=seq(1:5))
levels(df$letters)
subdf <- df %>% subset(numbers <= 3)
levels(subdf$letters)
subdf$letters <- droplevels(subdf$letters)
levels(subdf$letters)
```

### Replace NA with value in factor

Replace NA with value in a factor variables.

```{r}
replaceFactorNAs <- function(fctr, naLevel = "Unknown"){
  ifelse(fctr %>% is.na, naLevel, fctr %>% as.character) %>%
    factor(fctr %>% levels %>% append(naLevel))
}
```

### getFactors

extract factor columns from data frame.

```{r}
library(magrittr)
getFactors <- function(dataFrame){
  dataFrame %>%
    names %>%
    sapply(function(colName){
             dataFrame %>%
               getElement(colName) %>%
               is.factor
    }) %>%
    dataFrame[,.]
}
exams <-
  expand.grid(res = c("Y", "N"),
              score = 1:4,
              id = LETTERS[1:5])
exams %>% head %T>% print %>% getFactors
```

### ordered factors

with ordered / ordinal factors you can filter using less / greater / larger / smaller than.

```{r "ordered factors"}
mp <- structure(list(pa = c("M 1", "M 2", "M 3")
                     ,fy = c("FY17.05", "FY17.04", "FY17.03"))
                , .Names = c("pa", "fy")
                , row.names = c(NA, -3L)
                , class = "data.frame")
mp %<>%
  dplyr::mutate( pa = pa %>% ordered
                ,fy = fy %>% as.ordered)
mp %>% str
mp %>% dplyr::filter(pa > "M 1")
```

### transforming factors

```{r}
library(magrittr)
levels <- 1:4
factorVariable <- factor(levels,
                         labels=levels+10)
factorVariable %>% as.integer # returns underlying levels
factorVariable %>% as.character # returns labels
```

### compact factor

to weed out empty factors

```{r "compact factor"}
asCompactFactor <- function(fctr) fctr %>% as.character %>% as.factor
(nonCompactFactor <- factor(c("A", "C", "A"), levels = LETTERS[1:5]))
nonCompactFactor %>% asCompactFactor

```

### edit factors

alter / remap factor-levels

- use a look-up table
- `remapFactor` function is MUCH faster for large factors
- NOTE: the mapping is reversed in the two approaches
- NOTE: use `as.character` in remapFactor2 to avoid serious bugs

remapFactor code taken from [vignette](https://cran.r-project.org/web/packages/gdata/vignettes/mapLevels.pdf)

```{r}
(fvar <-
  gl(n = 3, # number of levels
     k = 2, # number of replications
     labels = c("Control", "Treatment1", "Treatment2")))

remapFactor <- function(fctr, fctrMap){ levels(fctr) <- fctrMap ; fctr }
remapFactor2 <- function(fctr, fctrMap){ fctr %>% as.character %>% fctrMap[.] %>% unname %>% as.factor }

factorLookUpMap <-
  list(
        "trmt" = "Treatment1"
       ,"trmt" = "Treatment2"
       ,"ctrl" = "Control"
       )
fvar %>% remapFactor(factorLookUpMap)

factorLookUpMap2 <-
  c(
     "Treatment1" = "trmt"
    ,"Treatment2" = "trmt"
    ,"Control"    = "ctrl"
    )
fvar %>% remapFactor2(factorLookUpMap2)

```

or use `factor` to reorder factor levels

```{r}
set.seed(12)
ist <- InsectSprays$spray %>% sample %>% head(10) %>% as.character
ist %>% factor(levels = c("F", "D", "E", "A", "B"))
.Last.value %>% table
```

or when it's more complicated

```{r}
c(1, 0, 0, NA, 1) %>%
  factor(labels = c("N", "Y")) %>%
  factor(c("N", "Y", "U"))
```

### rm one level factors

from a data frame you might want to remove columns that are factors with only one level (since there's no variance). (compact factor / sparse factor)

```{r "rmOneLevelFactors"}
isOneLevelFactor <- function(vctr) {(vctr %>% is.factor) & (vctr %>% levels %>% length %>% equals(1))}
rmOneLevelFactors <- function(dta)
  dta %>%
    purrr::map(function(col) if(col %>% isOneLevelFactor) NULL else col ) %>%
    compact %>% as.data.frame

data.frame(a = gl(2,2), b = gl(1,4), d = 1:4) %>% rmOneLevelFactors

```

