### Overview

section for working with lists

### Subsetting

Subset / Select specific elements in a list.

Use single square brackets and use numerical index >.<

TODO: Add Helper for this??

```{r "list select"}
Harman23.cor[c(2,3)] %>% str
```

using names to exclude / include list elements

```{r "named list subsetting"}
include.list <- function(vctr, elements) vctr[vctr %>% names %in% elements]
exclude.list <- function(vctr, elements) vctr[! vctr %>% names %in% elements]
l1 <-
  list(
     qsec = 2
    ,mpg = 3
    ,disp = 4
    ,vs = 5
    ,wt = 6
    ,am = 7
   )

l1 %>% include(c("qsec", "vs", "am")) %>% str
l1 %>% exclude(c("qsec", "vs", "am")) %>% str
```

### combining

adding / appending / pushing / concatenating lists can be done with `append`

```{r "combing lists"}

la <- list(a = 1:2, rr = runif(4))
lb <- list(b = 3)
append(la, lb)

```

same with tibble

```{r}

library(tibble)
la <- lst(a = 1:2, rr = runif(4))
lb <- lst(b = 3)
append(la, lb)

```

lists with overlapping object names will still merge (not overwrite) and extraction gets iffy

```{r "combining list gotcha"}

la <- list(a = 1:2, rr = runif(4))
lr <- list(rr = 3)
(ll <- append(la, lr))
# grabs the first object with corresponding name. use index to access the later "rr"
ll[["rr"]]

```

tibble is consistent with list

```{r}

la <- tibble::lst(a = 1:2, rr = runif(4))
lr <- tibble::lst(rr = 3)
(ll <- append(la, lr))
ll[["rr"]]

purrr::update_list(la, !!! lr) # use purrr::update_list with splicing !!! / UQS()

```



### multi-layer lists

build larger lists and iterate with them

```{r "multi-layer lists"}

library(magrittr)
dates <-
  list( day1 = "2015-12-12"
       ,day2 = "2016-07-12"
       ,day3 = "2014-01-03") %>%
    sapply(as.Date, simplify = FALSE)
wkdays <- dates %>% sapply(weekdays, simplify = FALSE)
mnths <- dates %>% sapply(months, simplify = FALSE)

```

How do we combine these in one named list?

```{r}

mapply(FUN = function(dta, name){
         list( dta = dta
              ,weekday = wkdays[[name]]
              ,month = mnths[[name]]
              )
       }
  ,dates
  ,names(dates)
  ,SIMPLIFY = FALSE
  )

```

or even better

```{r}

dd <- dates %>% sapply(function(dta){list(date = dta)}, simplify = FALSE)
dd %>%
  sapply(function(lst){
           append(lst, list(weekday = lst$date %>% weekdays))
          }, simplify = FALSE)

```

or more succinct if it makes sense

```{r}

dates %>%
  sapply(function(dta){
           list( date = dta
                ,weekday = dta %>% weekdays)
          }, simplify = FALSE)

```

use tibble and purrr

```{r "multi-layers with purrr and tibble"}
library(tibble)
library(purrr)
dates %>%
  as.tibble %>%
  map(function(date){
        lst( date = date
            ,weekday = date %>% weekdays)
       }) # %>% as.tibble

```

multiple arguments to map does NOT work as expected

```{r}

dates %>%
  as.tibble %>%
  map(dates %>% names
      ,.f = function(date, dayLabel){
        lst( date = date
            ,month = date %>% months
            ,label = dayLabel)
      })

```

mix nested lists and vectors

```{r}
datesList <-
  dates %>%
    map(function(date){
      lst( date = date
          ,meta = lst( weekday = date %>% weekdays)
         )
     })

datesList %>% map(function(ddt){
    ddt$meta %<>% append(list(month = ddt$date %>% months))
    ddt
   })

```

### iteration and names

preserving names in the returning list

```{r "lists and preserving names"}
library(magrittr)
dayList <- list(Day1 = 12, Day2 = 34, `Day 3` = 23)
sapply(dayList, sqrt, simplify = FALSE)
```

accessing list item name inside function

```{r "list item access from inside function"}

dayPrinter <- function(dta, listName) paste0(listName, ": ", dta)
Map(dayPrinter
    ,dayList
    ,names(dayList))

```

or with `mapply`, since `Map` is directly using `mapply`

```{r}

mapply(FUN = dayPrinter, dayList, names(dayList), SIMPLIFY = FALSE)

```

use a function to remove the boiler plate

```{r}

mapplyWithNames <- function(lst, FUN) mapply(FUN = FUN, lst, names(lst), SIMPLIFY = FALSE)
dayList %>% mapplyWithNames(dayPrinter)

```


