

Classification in Kyphosis dataset like we did with `rpart`

```{r "party::ctree on Kyphosis data"}
library(rpart)
library(partykit) # library(party)
ctree(Kyphosis ~ Age + Number + Start
      ,data = kyphosis) %T>% print %>% plot(main = "Conditional Inference of Kyphosis data")
```

regression

```{r "party::ctree regression"}
set.seed(290875)

ozoneDt <- airquality %>% dplyr::filter(!is.na(Ozone))
airct <-
  ctree( Ozone ~ .
        ,data = ozoneDt)
airct %T>% print %>% plot
```

you can grab the data in the different segments clusters using `party::where`. using this you can extract cluster data or aggregates.

```{r "party::ctree segments data/aggregates"}
ozoneDt %>%
  dplyr::mutate(segment = predict(airct)) %>%
  dplyr::group_by(segment) %>%
  dplyr::summarise(avgOzone = mean(Ozone, na.rm = TRUE))

```

Mean squared error

```{r, eval = FALSE}
mean((airq$Ozone - predict(airct))^2)
```

extract observation terminal node ID using either `predict` or `where`

```{r "party::ctree regression data extraction"}
airct %>% predict(type = "node") # airct %>% party::where # only in party
```

```{r "party::ctree BinaryTree handling", eval = FALSE}
airct %>% nodes(4) # extract nodes # party only
airct %>% response %>% summary # extract the response variable from the formula. # party only
```

nominal variable classification

```{r "party::ctree classification"}
irisct <- ctree(Species ~ .,data = iris)
irisct %T>% print %>% plot
table(predict(irisct), iris$Species)
```

response of tree for the given observations (probabilities of levels in Species)

```{r, eval = FALSE}
treeresponse(irisct, newdata = iris[c(1,61, 120),]) # party only
```

ordinal regression. `ME` from the `mammoexp` data frame is an ordinal factor.

```{r "party::ctree ordinal regression"}
data("mammoexp", package = "TH.data")
mammoct <- ctree(ME ~ ., data = mammoexp)
plot(mammoct)
```

response of tree for the given observations (probabilities of levels in ME)

```{r, eval = FALSE}
treeresponse(mammoct, newdata = mammoexp[1:4,]) # party only
```

survival analysis.

```{r "party::ctree survival"}
data("GBSG2", package = "TH.data")
GBSG2 %>% str
GBSG2ct <- ctree(survival::Surv(time, cens) ~ .,data = GBSG2)
plot(GBSG2ct)

```

the tree response here is a whole survival fit (survfit object, e.g. can be plotted)

```{r, eval = FALSE}
treeresponse(GBSG2ct, newdata = GBSG2[1:2,]) # party only
```

options (see `?ctree_control` for options)

- `mincriterion` for significance level
- `testtype` for multiple testing method
- `minbucket` for smallest possiple bucket
- `maxsurrogate` unsure!

```{r}
data(plantTraits, package = "cluster")
ctree( height ~ .
      ,data = plantTraits
      ,mincriterion = .9
      # ,testtype = "Teststatistic"
      ) %>% plot

ctree( height ~ .
      ,data = plantTraits
      ,testtype = "Bonferroni"
      ) %>% plot

ctree( height ~ .
      ,data = plantTraits
      ,mincriterion = .8
      ,testtype = "Bonferroni"
      ) %>% plot

ctree( height ~ .
      ,data = plantTraits
      ,minbucket = 10
      # ,testtype = "Teststatistic"
      ) %>% plot

ctree( height ~ .
      ,data = plantTraits
      ,maxdepth = 3 #,testtype = "Teststatistic"
      ) %>% plot

ctree( height ~ .
      ,data = plantTraits
      ,mincriterion = .9 #,testtype = "Teststatistic"
      ) %>% plot

ctree( height ~ .
      ,data = plantTraits
      ,mincriterion = .9 ,maxsurrogate = 2 # ,testtype = "teststatistic"
      ) %>% plot
```

extracting node splits conditions for each terminal node.

[SO link](https://stackoverflow.com/questions/21443203/ctree-how-to-get-the-list-of-splitting-conditions-for-each-terminal-node)

```{r "install emilrehnberg/party.readpaths", eval = FALSE}
devtools::install_github("emilrehnberg/party.readpaths")
```

```{r "generated data and ctree"}
shiftFirstPart <- function(vctr, divideBy, proportion = .5){
  vctr[vctr %>% length %>% multiply_by(proportion) %>% round %>% seq] %<>% divide_by(divideBy)
  vctr
}
set.seed(11)
n <- 13000
gdt <-
  data.frame( is_buyer = runif(n) %>% shiftFirstPart(1.5) %>% round %>% factor(labels = c("no", "yes"))
             ,age = runif(n) %>% shiftFirstPart(1.5) %>%
               cut(breaks = c(0, .3, .6, 1), include_lowest = TRUE, ordered_result = TRUE, labels = c("low", "mid", "high"))
             ,city = runif(n) %>% shiftFirstPart(1.5) %>%
               cut(breaks = seq(0,1,.2), include_lowest = TRUE, labels = c("Chigaco", "Boston", "Memphis", "LA", "ATL"))
             ,point = runif(n) %>% shiftFirstPart(1.2)
             )
gct <- party::ctree( is_buyer ~ ., data = gdt)

party.readpaths::readCtreePaths(gct, gdt) # dts <- gdt; ct <- gct
gct %>% plot

```

In `partykit`

```{r, eval = FALSE}
partykit:::.list.rules.party(ct)
```

```{r}
detach(package:partykit)
airq <- subset(airquality, !is.na(Ozone))
act <-
  party::ctree( Ozone ~ .
        ,data = airq # ,maxsurrogate = 3 # partykit
        ,controls = party::ctree_control(maxsurrogate = 3) # party
        )
party.readpaths::readCtreePaths(act, airq) # party only

data(plantTraits, package = "cluster")
pct <- party::ctree( height ~ .  ,data = plantTraits)
party.readpaths::readCtreePaths(pct, plantTraits)
```

prediction for the `party` package.

terminal nodes & response data prediction

```{r "prediction on party::ctree objects"}
pct %>% party::where(newdata = plantTraits[1:10,])
pct %>% party::treeresponse(newdata = plantTraits[1:10,]) # not sure what this is.. >.<
```

response prediction

```{r}
pct %>% predict(newdata = plantTraits[1:10,])
```


