
## Overview

Section for statistics. Distributions, testing etc.

## quantiles

quantiles for different probabilities, use `stats:quantile`

```{r "quantile"}

randomNormals <- rnorm(1001)
randomNormals %>% quantile
randomNormals %>% quantile(.4)
randomNormals %>% quantile(probs = c(.1, .5, 1, 2, 5, 10, 50, NA)/ 100)

```

## Distributions

### Overview

Section for probability distributions, e.g. Normal/Gaussian, Binomial, Beta

### Shapes

Randomize from distributions and introduce shapes.

Skewness

```{r "skewness-randomize"}
```

Kurtosis

```{r "kurtosis-randomize"}
```

### Beta distribution

- Typically used to model probabilities.
- Conjugate Prior to a Binomial distribution.
- Say that we are modelling a baseball batter
  - alpha parameter can be interpreted as number of hits
  - beta parameter can be interpreted as number of misses
- Ref: [wiki](https://en.wikipedia.org/wiki/Beta_distribution)

Let's look at a Beta distribution with alpha = 81 and beta = 219.

```{r}

alpha0 <- 81
beta0 <- 219

qbeta(.5, alpha0, beta0)
alpha0 / (alpha0 + beta0) # mean for this probability

```

standard beta distribution function. shape1/2 being alpha/beta.

```{r, eval = FALSE}
dbeta(x, shape1, shape2, ncp = 0, log = FALSE)
pbeta(q, shape1, shape2, ncp = 0, lower.tail = TRUE, log.p = FALSE)
qbeta(p, shape1, shape2, ncp = 0, lower.tail = TRUE, log.p = FALSE)
rbeta(n, shape1, shape2, ncp = 0)
```

- `dbeta` - returns density. high values for `x` close to the mean. E.g. usable for plotting the distribution.
- `qbeta` - returns quantile. E.g. 0.5 quantile is the mean.
- `pbeta` - returns probability. E.g. the probability of hitting 0.25 or less is 22% but ~88% to be under 0.33 (obviously this is pure theory for a probability in a one event trial)
- `rbeta` - returns random variables outcomes from a Beta(alpha0, beta0)

```{r}

n <- 21
x <- seq(0, 1, length = n)
data.frame(x
           ,d = dbeta(x, alpha0, beta0)
           ,quantile = qbeta(x, alpha0, beta0) # cumulative quantile of attaning probability x
           ,prob = pbeta(x, alpha0, beta0) # cumulative probability of attaning quantile x
           ,random = rbeta(n, alpha0, beta0) # random outcomes of n beta distributions with alpha0/beta0
           ) %>% round(3)

```

Bayesian setting. New data comes in.

- The batter has 100 more hits and out of 300 tries.
- update in bayesian way using bayes rule
  - Likelihood follows a Binomial distribution.
  - The posterior is another Beta with new alpha and beta.

```{r}

alpha1 <- alpha0 + 100 # add the hits
beta1 <- beta0 + 300 - 100 # add the misses

data.frame(x
           ,q0 = qbeta(x, alpha0, beta0)
           ,q1 = qbeta(x, alpha1, beta1)
           )

```

Plotting this distributions

```{r}

library(ggplot2)

sim <- data.frame(a = c(81, 81 + 100),
                  b = c(219, 219 + 200)) %>%
  dplyr::group_by(a, b) %>%
  dplyr::do(dplyr::data_frame(x = seq(0, 1, .001), y = dbeta(x, .$a, .$b))) %>%
  dplyr::mutate(Parameters = paste0("\u03B1 = ", a, ", \u03B2 = ", b)) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(Parameters = factor(Parameters, levels = unique(Parameters)))

sim %>% dplyr::filter(a == 81) %>%
  ggplot(aes(x, y, color = Parameters)) + geom_line() +
  xlim(0, .5) + ylab("Density of beta")

sim %>%
  ggplot(aes(x, y, color = Parameters)) + geom_line() +
  xlim(0, .5) + ylab("Density of beta")

```


## Statistical Tests

### Overview

Section for test. Parametric and non-paratmetric.

### Mann-Whitney-Wilcoxon Test

In statistics, the Mann-Whitney U test (also called the Mann-Whitney-Wilcoxon (MWW), Wilcoxon rank-sum test, or Wilcoxon-Mann–Whitney test) is a nonparametric test of the null hypothesis that two samples come from the same population against an alternative hypothesis, especially that a particular population tends to have larger values than the other.[@wikiMannWhitneyUTest]

Unlike the t-test it does not require the assumption of normal distributions. It is nearly as efficient as the t-test on normal distributions.[@wikiMannWhitneyUTest]

Use the Wilcoxon signed-rank test when samples are related / paired.

Example code [@wilcoxTest]

Two-sample test.

```{r}
library(graphics)
x <- c(0.80, 0.83, 1.89, 1.04, 1.45, 1.38, 1.91, 1.64, 0.73, 1.46)
y <- c(1.15, 0.88, 0.90, 0.74, 1.21)
wilcox.test(x, y, alternative = "greater")
wilcox.test(x, y, alternative = "greater",
            exact = FALSE, correct = FALSE) # H&W large sample approximation

wilcox.test(rnorm(10), rnorm(10, 2), conf.int = TRUE)

```

Formula interface.

```{r}

boxplot(Ozone ~ Month, data = airquality)
wilcox.test(Ozone ~ Month, data = airquality,
            subset = Month %in% c(5, 8))

```

### Wilcoxon signed-rank test

The Wilcoxon signed-rank test is a non-parametric statistical hypothesis test used when comparing two related samples, matched samples, or repeated measurements on a single sample to assess whether their population mean ranks differ (i.e. it is a paired difference test). It can be used as an alternative to the paired Student's t-test, t-test for matched pairs, or the t-test for dependent samples when the population cannot be assumed to be normally distributed. [@lowry11]

Example code [@wilcoxTest], [@rTutorWilcoxonSignedRankTest]

```{r}
x <- c(1.83,  0.50,  1.62,  2.48, 1.68, 1.88, 1.55, 3.06, 1.30)
y <- c(0.878, 0.647, 0.598, 2.05, 1.06, 1.29, 1.06, 3.14, 1.29)

wilcox.test(x, y, paired = TRUE, alternative = "greater")
```

One-sample test version.

```{r }
wilcox.test(y - x, alternative = "less",
            exact = FALSE, correct = FALSE) # H&W large sample approximation

library(MASS)
wilcox.test(immer$Y1, immer$Y2, paired=TRUE)
```

### Contingency table test

look into the `stats` cheatsheet for usage of these tests.

example data

```{r "data for contingency tables test"}
library(magrittr)
partyGenderTable <-
  structure( c(762, 484, 327, 239, 468, 477)
            ,.Dim = 2:3
            ,.Dimnames = structure(list( gender = c("F", "M")
                                        ,party = c("Democrat", "Independent", "Republican"))
            ,.Names = c("gender", "party"))
            ,class = "table")

set.seed(12)
experimentDt <-
  data.frame( exposure = gl(2, 20, labels = c("Control", "Treatment"))
             ,outcome = gl(2, 20, labels = c("Sick", "Healhty")) %>% sample(replace = TRUE)
             )
```

#### Pearson chi-square test

can be applied to tables.

```{r}
(XsqTest <- partyGenderTable %>% chisq.test)
XsqTest$observed   # observed counts (same as M)
XsqTest$expected   # expected counts under the null
XsqTest$residuals  # Pearson residuals
XsqTest$stdres     # standardized residuals
```

or to variables directly.

```{r}
with(experimentDt, { chisq.test(exposure, outcome) })
```

#### Fisher's exact test

use function `fisher.test` and apply similar to `chisq.test`.

#### McNemar's test

for paired observations. use function `mcnemar.test`.

```{r "McNemars test"}
(Performance <-
  matrix(c(794, 86, 150, 570),
         nrow = 2,
         dimnames = list("1st Survey" = c("Approve", "Disapprove"),
                         "2nd Survey" = c("Approve", "Disapprove")))
 )

Performance %>% mcnemar.test
```

Here there's a significant difference in the approval from the 1st to the 2nd survey.

### ANOVA

```{r "ANOVA and Tukey HSD pairwise comparisons"}
library(magrittr)
warpbreaks %T>%
  { message("Data Summary"); print(summary(.)); message() } %>%
  aov(breaks ~ wool + tension, data = .) %T>%
  { message("ANOVA Summary"); print(summary(.)); message(); message("Pairwise comparisons") } %>%
  TukeyHSD("tension", ordered = TRUE)
```

## Supervised

```{r Supervised, child="supervised.Rmd", cache=FALSE}
```

## Unsupervised

```{r Unsupervised, child="unsupervised.Rmd", cache=FALSE}
```

## Forecasting

```{r, echo=FALSE}
knitr::read_chunk("../package-demos/forecast.R")
```

```{r forecast-pkg-demo}
```

## Neural Nets

```{r NNs, child="neural-nets.Rmd", cache=FALSE, eval=TRUE}
```
