
## Overview

Section for "tricks".

## saving files

`save` saves the R-object(s) to disc and keeps the naming

```{r, eval = FALSE}
save(foo,file="data.Rda")
load("data.Rda")
```

`saveRDS` saves one object and does not keep the name for the object.

```{r, eval = FALSE}
saveRDS(foo, file="data.Rda") # even for S4 objects
bar <- readRDS(file="data.Rda")
```

```{r, eval = FALSE}
write
write.table
write.csv
```

```{r, eval = FALSE}
library(MASS)
mat <- matrix(1:100,nrow=20)
write.matrix(mat,'file.prn',sep = "\t")

mat2 <- as.matrix(read.table("file.prn", as.is = TRUE))
mat2 <- unname(mat2) # make mat2 a true matrix by removing the names
all.equal(mat, mat2) # [1] TRUE
```

```{r, eval = FALSE}
dput; dget # not for S4 objects
```

```{r, eval = FALSE}
dump; source # not for S4 objects
```

## remove names

```{r}
require(magrittr)
dt <- data.frame(V1 = runif(100) %>% round,
                 V2 = runif(100) %>% round)
(crossMatrix <-
  ( with(dt, table(V1,V2))
    %>% as.data.frame(row.names = LETTERS[1:4])
    %>% as.matrix
    )
  )
crossMatrix %>% unname
```

or NULLify either row or column names

```{r}

colnames(crossMatrix) <- NULL
crossMatrix
rownames(crossMatrix) <- NULL
crossMatrix

```

## remove named columns

```{r message=FALSE}
require(magrittr)
```

```{r}
df <- data.frame(x=1:5, y=2:6, z=3:7, u=4:8)
df
df %>% subset(select=-c(z,u))
df[ , -which(names(df) %in% c("z","u"))] # also works but not as nice IMO
```

## insertion in characters

or strings. how to find something similar to the use case of Ruby's `insert`

```{r "character insertion"}
"shjat" %>% gsub('^([a-z]{2})([a-z]+)$',"\\1_\\2",.)

insert <- function(obj, ...) UseMethod("insert")
insert.character <- function(chr, position, insertion){ # only works with \w characters
  chr %>%
    gsub( paste0("^(\\w{", position - 1, "})(\\w+)$")
         ,paste0("\\1", insertion, "\\2")
         ,.)
}
"shjat" %>% insert(3, "_")
```

## time-series convertions

Dependency installation

```{r, eval = FALSE}

dependencies <- c("zoo", "xts")
devtools::install_github(paste("cran", dependencies, sep = "/"))

```


```{r}

library(xts)
price <- structure(list(
  date = c(20070103L, 20070104L, 20070105L, 20070108L, 20070109L,
           20070110L, 20070111L, 20070112L, 20070115L),
  close = c(54.7, 54.77, 55.12, 54.87, 54.86, 54.27, 54.77, 55.36, 55.76)),
  .Names = c("date", "close"), class = "data.frame",
  row.names = c("1", "2", "3", "4", "5", "6", "7", "8", "9"))
price$date %<>% as.character %>% as.Date(format="%Y%m%d")
xts(price$close, price$date)

```

```{r}

library(zoo)
pricez <- read.zoo(text="   DATE  CLOSE
 1    20070103 54.700
 2    20070104 54.770
 3    20070105 55.120
 4    20070108 54.870
 5    20070109 54.860
 6    20070110 54.270
 7    20070111 54.770
 8    20070112 55.360
 9    20070115 55.760
 ")
index(pricez) <- as.Date(as.character(index(pricez)), format="%Y%m%d")
pricez

```

`ts` (time-series) object to `data.frame`

```{r}
library(zoo)
UKgas %>%
  { data.frame(date = as.Date(as.yearqtr(time(.))),
               gas = as.matrix(.)) }
```

## Tables

```{r, eval=FALSE, echo=FALSE}

dependencies <- c("gtools", "gdata", "gmodels")
devtools::install_github(paste("cran", dependencies, sep = "/"))

```

```{r}

library(gmodels)
data(infert, package = "datasets")

CrossTable(warpbreaks$wool,
           warpbreaks$tension,
           chisq = TRUE,
           prop.t = TRUE,
           digits = 2,
           dnn = c("Wool", "Tension"))

CrossTable(infert$education, infert$induced, expected = TRUE)
CrossTable(infert$education, infert$induced, expected = TRUE, format="SAS")
CrossTable(infert$education, infert$induced, expected = TRUE, format="SPSS")
```

## cumulatively sum within a group

Cumulativily sum the `carb` column within each `am`-group

```{r}
library(magrittr)
library(dplyr)
( mtcars
  %>% filter(cyl == "6")
  %>% select(carb, gear, am)
  %>% group_by(am)
  %>% arrange(am)
  %T>% print
  %>% do(cumsum(.["carb"]))
  )
```

## list to data.frame

given a list of data frames. use `Reduce` to bind them togetherq

```{r}
library(magrittr)
library(dplyr)
c(3, 6, 12) %>%
  lapply(function(day){ airquality %>% filter(Day == day) }) %>%
  Reduce(rbind, .)
```

