## Overview

Section for good helper functions.

## exclude

delete / remove / exclude / reject elements from a vector

```{r "vector remove element(s)"}
exclude.default <- function(vctr, elements) vctr[! vctr %in% elements]
a <- c("master", "of", "puppets")
a %>% exclude(c("of", "master"))
```

## Curry

perform currying in R with the `functional` package. Or just use

```{r}
Curry <- function(FUN, ...){
  .orig <- list(...)
  function(...) do.call(FUN, c(.orig, list(...)))
}
```

## isEmpty

checks if list / atomic vector (array) is empty

```{r}
library(magrittr)
isEmpty <- function(obj) UseMethod("isEmpty")
isEmpty.default <- function(obj) obj %>% length %>% equals(0)
isEmpty.data.frame <- function(dtf) dtf %>% nrow %>% equals(0)
isEmpty.tbl <- function(obj) obj %>% as.data.frame %>% isEmpty
isEmpty.character <- function(chrs) chrs %>% `==`("")

integer(0) %>% isEmpty
1:3 %>% isEmpty
list() %>% isEmpty
1:3 %>% as.list %>% isEmpty
data.frame(price = integer(0)) %>% isEmpty
data.frame(price = 1:3) %>% isEmpty
"" %>% isEmpty
'' %>% isEmpty
'a' %>% isEmpty
```

## isNonEmpty

checks if list / atomic vector (array) is non-empty

```{r}
library(magrittr)
isNonEmpty <- function(obj) UseMethod("isNonEmpty")
isNonEmpty.default <- function(obj) obj %>% length %>% is_greater_than(0)
isNonEmpty.data.frame <- function(obj) obj %>% nrow %>% is_greater_than(0)
isNonEmpty.tbl <- function(obj) obj %>% as.data.frame %>% isNonEmpty

integer(0) %>% isNonEmpty
1:3 %>% isNonEmpty
list() %>% isNonEmpty
1:3 %>% as.list %>% isNonEmpty
data.frame(price = integer(0)) %>% isNonEmpty
data.frame(price = 1:3) %>% isNonEmpty
```

## asQuoted

put single quotes around the argument. This is useful mainly for the underscore variants of `dplyr` functions.

```{r "asQuoted"}
asQuoted <- function(arg) paste0("'", arg, "'")
set.seed(12)
threshold <- .7
data.frame(a = runif(10)) %>% filter_(paste("a > ", threshold %>% asQuoted))
```

## compact

similar to ruby's `compact`.

```{r}

library(magrittr)
compact <- function(object) { UseMethod("compact") }
compact.default <- function(obj) obj[which(!is.na(obj))]
compact.character <- function(obj) obj[which(obj != "")] %>% compact.default
compact.list <- function(lst){
  lst[lapply(lst,
             function(elm){
               if(elm %>% is.atomic){
                 elm %>% length
               }else{
                 elm %>% nrow
               }
             }) > 0]
}

( list(a = 1,
       b = 2,
       c = integer(0),
       d = character(0),
       e = 3,
       f = data.frame(ii = integer(0), cc = character(0))
       ,g = NA
       ) %>% compact
  )

```

## prepend

add `prepend` (for nice naming) since `append` exists

```{r}
prepend <- function(x, values) { append(values, x) }
append(1:2, 6:8)
prepend(1:2, 6:8)
```

## append / prepend class

```{r "prepend class"}

appendClass <- function(obj, cls) {
  class(obj) %<>% append(cls)
  obj
}

prependClass <- function(obj, cls) {
  class(obj) %<>% prepend(cls)
  obj
}

```

## wordArray

inspired by ruby's `%w`

```{r}
library(magrittr)
wordArray <- function(string){ string %>% strsplit("[[:space:]]+") %>% unlist }
"carat color value" %>% wordArray
```

## dataDivision

Divide by zero with set default

```{r}
library(magrittr)
dataDivision <- function(numerator, denominator, divZero = NA){
  ( numerator / denominator ) %>%
   { ifelse(is.finite(.), ., divZero) }
}
dataDivision(10,2)
dataDivision(10,0)
```

## Diagonal proportion of a matrix

returns if object is square shaped i.e. number of rows and columns are the equal

```{r}
isSquareShaped <- function(object) { equals(ncol(object), nrow(object)) }
```

calculates the diagonal proportion in a square-shaped object

```{r}
library(magrittr)
diagonalProportion <-
  function(object){
    if(object %>% isSquareShaped %>% not){
      warning("calculating the diagonal proportion only applies to squared objects")
      return(0)
    }
    divide_by(object %>% diag %>% sum,
              object %>% sum)}
(diag(3) + 0.5) %T>% print %>% diagonalProportion
```

## Interval

function for creating an interval sequence (integers)

```{r}
library(magrittr)
interval <- function(mid, radius) { mid %>% { seq(. - radius, . + radius) } }
```

## mkdir -p

function for creating deep folder seemlessly

```{r}
library(magrittr)
mkdirp <- function(filepath) { filepath %>% dir.create(recursive = TRUE, showWarnings = FALSE) }
```

## rest

function similar to `rest`/`next`/`pop` in closure. I.e. returns object without the first element.
works with e.g. atomic vectors and lists. (will not work with a `data.frame`)

```{r}
library(magrittr)
rest <- function(stk){
  if(stk %>% length %>% is_greater_than(1) ){
    stk[2:length(stk)]
  }else{
    NULL
  }
}

1:4 %>% rest
1:4 %>% as.list %>% rest
```

## truncateLast

similar to the above `rest` function but for the last object in a stack
works with e.g. atomic vectors and lists. (will not work with a `data.frame`)

```{r}
library(magrittr)
truncateLast <- function(stk){
  if(stk %>% length %>% is_greater_than(1) ){
    stk[1:(length(stk) - 1)]
  }else{
    NULL
  }
}

4:1 %>% truncateLast
4:1 %>% as.list %>% truncateLast
```

## coalesce function

Using `Reduce`, I don't know *why* this works >.<

```{r}
a <- c(1,  2,  NA, 4, NA)
b <- c(NA, NA, NA, 5, 6)
c <- c(7,  8,  NA, 9, 10)

coalesce <- function(...) {
  list(...) %>%
    Reduce(f = function(nxt, acc) {
      i <- which(is.na(nxt))
      nxt[i] <- acc[i]
      nxt})
}
coalesce(a, b, c)
```

based on [stack overflow](http://stackoverflow.com/questions/19253820/how-to-implement-coalesce-efficiently-in-r) the manual index version below is faster.

```{r}
coalesce <- function(...) {
  output <- ..1
  for (element in list(...)[-1]) {
    naIndexes <- which(is.na(output))
    output[naIndexes] <- element[naIndexes]
  }
  output
}
coalesce(a, b, c)
```

## multiple plots

using `ggplot` get multiple plots in the same window

TODO: add to package, would be nice, actually maybe make a helper package w documentation etc.

Multiple plot function

ggplot objects can be passed in `...`, or to plotlist (as a list of ggplot objects)
- `cols`:   Number of columns in layout
- `layout`: A matrix specifying the layout. If present, 'cols' is ignored.

If the layout is something like `matrix(c(1,2,3,3), nrow=2, byrow=TRUE)`,
then plot 1 will go in the upper left, 2 will go in the upper right, and
3 will go all the way across the bottom.


```{r "multiplot function"}

multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```

`multiplot` demo

```{r}
library(ggplot2)

p1 <- ggplot(ChickWeight, aes(x=Time, y=weight, colour=Diet, group=Chick)) +
	geom_line() +
	ggtitle("Growth curve for individual chicks")

p2 <- ggplot(ChickWeight, aes(x=Time, y=weight, colour=Diet)) +
	geom_point(alpha=.3) +
	geom_smooth(alpha=.2, size=1) +
	ggtitle("Fitted growth curve per diet")

p3 <- ggplot(subset(ChickWeight, Time==21), aes(x=weight, colour=Diet)) +
	geom_density() +
	ggtitle("Final weight, by diet")

p4 <- ggplot(subset(ChickWeight, Time==21), aes(x=weight, fill=Diet)) +
	geom_histogram(colour="black", binwidth=50) +
	facet_grid(Diet ~ .) +
	ggtitle("Final weight, by diet") +
	theme(legend.position="none") # No legend (redundant in this graph)

multiplot(p1, p2, p3, p4, cols = 2)
```

taken from [R cookbook](http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/)

