## Overview

notes for installing packages

## unable to load shared object

The `libcoin` package has some shared objects with the `mvtnorm` package.
So say that you install the `libcoin` package an get an error similar to `unable to load shared object: PATH/mvtnorm.so` appears during installation.

1. restart console
2. install the `mvtnorm` package including the `.so`
3. install the `libcoin` package

## after R upgrade

try to run `sudo R CMD javareconf` in the terminal.

```{r, eval = FALSE}
install.packages(c("devtools", "magrittr", "rJava", "dplyr"))
library(magrittr)
installFromCran <- function(pkgs) devtools::install_github(pkgs %>% paste("cran", ., sep = "/"))
partyPkgs <- c('mvtnorm', 'modeltools', 'zoo', 'sandwich', 'strucchange', 'TH.data', 'survival', 'multcomp', 'coin', 'party')
tibblePkgs <- c('rlang', 'tibble')
roxygen2Pkgs <- c('stringi', 'stringr', 'brew', 'backports', 'rprojroot', 'desc', 'commonmark', 'xml2', 'roxygen2')
testthatPkgs <- c('crayon', 'praise', 'testthat')
rattlePkgs <- c("RGtk2", "rattle")
lintrPkgs <- c( "registry", "iterators", "pkgmaker", "registry", "rngtools", "gridBase", "foreach", "doParallel"
               ,"NMF", "irlba", "rex", "stringdist", "igraph", "lintr")
rmdPkgs <- c("htmltools", "bitops", "caTools", "base64enc", "rmarkdown")
xlconnectPkgs <- c('rJava','XLConnectJars', 'XLConnect')
dplyrPkgs <- c( "bindr", "plogr", "bindrcpp", "glue", "pkgconfig", "rlang", "dplyr" )
rsqlPkgs <- c("dplyr", "DBI", "dbplyr", "RSQLServer")
knitrPkgs <- c("evaluate", "highr", "markdown", "yaml", "knitr")
c( partyPkgs, tibblePkgs, roxygen2Pkgs, testthatPkgs, rattlePkgs, lintrPkgs, rmdPkgs
  ,xlconnectPkgs, dplyrPkgs, rsqlPkgs, knitrPkgs, "tidyr", "matrixcalc", "dlm", "dbplyr") %>% installFromCran

devtools::install_github(c("emilrehnberg/party.readpaths"))
devtools::install_github(c("ramnathv/slidify"))
devtools::install_github(c("ramnathv/slidifyLibraries"))
```

No longer needed

```{r, eval = FALSE}
devtools::install_github(c( 'emilrehnberg/party.readpaths', 'jmp75/rClr'
                           ,"bescoto/RMSSQL", "bescoto/dplyr.mssql"))
```

copy below for running on server to install packages

```{r, eval = FALSE}
library(magrittr)
installFromCran <- function(pkgs) devtools::install_github(pkgs %>% paste("cran", ., sep = "/"))
dplyrPkgs <- c("bindr", "bindrcpp", "glue", "pkgconfig", "rlang", "dplyr" )
installFromCran(dplyrPkgs)
```

## tidyverse

```{r, "tidyverse installation", eval = FALSE}

c("r-dbi/DBI"
  ,"r-lib/rlang"
  ,"tidyverse/dplyr"
  ,"tidyverse/tibble"
  ,"tidyverse/purrr"
  ,"tidyverse/dbplyr"
  ,"cran/rematch" ,"cran/cellranger","tidyverse/readxl"
 ) %>% devtools::install_github()

c("r-lib/cli","r-lib/crayon"
  ,"rstudio/rstudioapi"
  ,"tidyverse/magrittr", "tidyverse/hms"
  ,"cran/BH", "mdlincoln/clipr" "tidyverse/readr"
  ,"tidyverse/forcats", "tidyverse/haven", "tidyverse/lubridate"
  ,"tidyverse/modelr"
  ,"r-lib/evaluate", "cran/yaml", "cran/markdown", "yihui/xfun", "yihui/highr", "yihui/knitr"
  ,"cran/base64enc", "eddelbuettel/digest", "cran/htmltools", "cran/jsonlite", "yihui/mime", "cran/tinytex","rstudio/rmarkdown"
  ,"r-lib/processx" , "r-lib/callr", "r-lib/fs", "edwindj/whisker", "r-lib/withr", "tidyverse/reprex"
  ) %>% devtools::install_github()

devtools::install_github( "tidyverse/tidyverse" )

```

### tibble

installing `tibble`

```{r, eval = FALSE}

depsLoc <- paste("r-lib", c("crayon", "pillar", "rlang"), sep = "/")
repoLoc <- "tidyverse/tibble"
c(depsLoc, repoLoc) %>% devtools::install_github()

```


### ggplot2

installation of `ggplot2` using `devtools`

```{r, "ggplot2 installation", eval = FALSE}

devtools::install_github(file.path("cran", c("RColorBrewer", "dichromat", "plyr", "colorspace", "munsell", "labeling", "viridisLite", "gtable", "lazyeval", "reshape2", "scales")))
devtools::install_github("tidyverse/ggplot2")

```

add-on packages

#### gifski

```{r "gifski installation", eval = FALSE}

depLoc <- "cran/png"
pkgLoc <- "r-rust/gifski"

c(depLoc, pkgLoc) %>% devtools::install_github(force = TRUE)

```


#### gganimate

```{r "gganimate installation", eval = FALSE}

tweenrLocs <- c("cran/farver", "cran/tweenr")
depsLoc <- c("r-lib/rlang", "r-lib/progress", "tidyverse/glue", tweenrLocs, "tidyverse/ggplot2")
pkgLoc <- "thomasp85/gganimate"
c(depsLoc, pkgLoc) %>% devtools::install_github()

```

#### plotly

```{r "plotly installation", eval = FALSE}

depLoc <-
  c("RcppCore/Rcpp", "rstudio/promises", "rstudio/httpuv", "rstudio/crosstalk", "Rdatatable/data.table", "cran/hexbin"
    ,"tidyverse/tibble", "r-lib/rlang", "tidyverse/purrr"
    ,"ramnathv/htmlwidgets", "r-lib/gtable", "tidyverse/ggplot2", "tidyverse/dplyr", "eddelbuettel/digest")
pkgLoc <- "ropensci/plotly"
c(depLoc, pkgLoc) %>% devtools::install_github(dependecies=TRUE)

```

## rvest

website scraper package. Needs `xml2` which looks for `xml2-config` and `pkg-config` both which has some shenanigans. Look into them and environment variables if you have issues.

```{r, "rvest installation", eval = FALSE}

c( "r-lib/xml2", "cran/httr", "hadley/rvest") %>% devtools::install_github()

```{r, eval = FALSE }
c("httpuv", "xtable", "sourcetools", "shiny") %>% paste0("cran/",.) %>% devtools::install_github()
```

## RODBC

and `RODBCext`. perform before trying to install on mac. look for other solutions on linux/mac.

```
brew update && brew install unixODBC
```

## party

install `party`.

```{r, eval = FALSE }
devtools::install_github('cran/party')
```

installing `party` dependencies.

```{r, eval = FALSE }
devtools::install_github('cran/mvtnorm')
devtools::install_github('cran/modeltools')
devtools::install_github('cran/zoo')      # sandwich dependency
devtools::install_github('cran/sandwich')
devtools::install_github('cran/strucchange')
devtools::install_github('cran/TH.data')  # multcomp dependency
devtools::install_github('cran/multcomp') # coin dependency
devtools::install_github('cran/coin')
devtools::install_github('cran/party')
```

## partykit

install `partykit`.

```{r, eval = FALSE }
devtools::install_github('cran/partykit')
```

installing `partykit` dependencies.

```{r, eval = FALSE }
devtools::install_github('cran/Formula')
```

## roxygen2

```{r, eval = FALSE }
devtools::install_github('cran/brew')
devtools::install_github('cran/backports') # rprojroot dependency
devtools::install_github('cran/rprojroot') # desc dependency
devtools::install_github('cran/desc')
devtools::install_github('cran/commonmark')
devtools::install_github('cran/xml2')
devtools::install_github('cran/roxygen2')
```

## testthat

```{r, eval = FALSE }

depsLoc <- c('r-lib/crayon', 'cran/praise')
pkgLoc <- 'r-lib/testthat'
c(depsLoc, pkgLoc) %>% devtools::install_github()

```

## rattle

```{r, eval = FALSE }
devtools::install_github('cran/rattle')
```

installing `rattle` dependencies. install `gtk+` via brew.

```{r, eval = FALSE }
devtools::install_github('cran/RGtk2')
```

## rJava

try to run `sudo R CMD javareconf` in the terminal.

```{r, eval=FALSE , echo=FALSE}
install.packages("rJava")
```

## lintr

installing lintr

```{r, eval = FALSE, echo = FALSE}
devtools::install_github("cran/registry") # pkgmaker dependency
devtools::install_github("cran/iterators") # foreach dependency
devtools::install_github( file.path("cran", c( "pkgmaker", "registry", "rngtools", "gridBase", "foreach", "doParallel" ))) # NMF dependencies
devtools::install_github( file.path("cran", c("NMF", "irlba"))) # igraph dependencies
devtools::install_github( file.path("cran", c("rex", "stringdist", "igraph")))
devtools::install_github("cran/lintr")

devtools::install_github( file.path("cran", c("rex", "stringdist", "igraph")))
devtools::install_github("cran/lintr")

```

## bestglm

installing `bestglm` package, similar to `leaps` but not limited to linear regression.

```{R "bestglm installation", eval = FALSE}

devtools::install_github(file.path("cran", c("glmnet", "grpreg")))
devtools::install_github("cran/bestglm")
library(bestglm)

```

## RSQLServer

install `rJava`.

CAREFUL! Do NOT install: `devtools::install_github('agstudy/rsqlserver')`

```{r, eval=FALSE, echo=FALSE }
install.packages("RSQLServer")
devtools::install_github('jmp75/rClr') # brew install mono # for this
devtools::install_github('imanuelcostigan/RSQLServer')
```

## dplyr.mssql

```{r, eval=FALSE , echo=FALSE}
devtools::install_github("bescoto/RMSSQL")
devtools::install_github("bescoto/dplyr.mssql") # this package
```

Also you'll need these packages: `DBI`, `dplyr`, `RJDBC`, `assertthat`, `dplyr`, `stringr`

If there's a timeout-problem, try changing download method.

```{r, eval=FALSE , echo=FALSE}
options(download.file.method = "wininet")

devtools::install_github("bescoto/dplyr.mssql", method = "wininet") # works? instead of setting the session options?
```

if there's problems with stringi, try below (the latter argument is for dealing with the case where the package has no matching version for R)

```{r, eval=FALSE , echo=FALSE}
R CMD INSTALL stringi_1.1.1.tar.gz --configure-args='--disable-pkg-config'
```

### Dependency hell

`dplyr`, `RSQLServer` and `dplyr.mssql` has this dependency hell going on ATM (161121). `dplyr.mssql` depends on `dplyr` version 4 but
latest `RSQLServer` requires >5.0. Downgrade `RSQLServer` and `dplyr` to make `dplyr.mssql` happy.

```{sh, eval=FALSE, echo=FALSE}
wget https://github.com/imanuelcostigan/RSQLServer/archive/v0.2.0.tar.gz
R CMD INSTALL v0.2.0.tar.gz

wget https://cran.rstudio.com/src/contrib/Archive/dplyr/dplyr_0.4.3.tar.gz
R CMD INSTALL dplyr_0.4.3.tar.gz

```

