---
title: "R cheatsheet"
author: Emil Rehnberg
date: "`r Sys.Date()`"
documentclass: krantz
biblio-style: apalike
link-citations: yes
colorlinks: yes
graphics: yes
lot: yes
lof: yes
fontsize: 11pt
mainfont: Palatino
monofontoptions: "Scale=0.8"
site: bookdown::bookdown_site
description: "Cheatsheet for R"
url: 'https\://github.com/EmilRehnberg/cheatsheets'
github-repo: EmilRehnberg/cheatsheets
output:
  pdf_document:
    highlight: zenburn
  html_document:
    toc: TRUE
    toc_depth: 3
    number_sections: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    css: [code.css, r.css, dark.css]
    df_print: tibble
---

```{r render, echo=FALSE, eval=FALSE}
rmarkdown::render("main.Rmd", output_format="html_document") # "all_document"
allexamples <- FALSE
```

# Set-up

Set-up section for this document.

Set options:

```{r set-options, echo=FALSE, cache=FALSE}
options(width = 200)
knitr::opts_chunk$set(tidy = "styler")
```

Read chunks:

```{r read-package-demos}
dir("demos", full.names = TRUE) %>% purrr::walk(knitr::read_chunk)
dir("demos/statistics", full.names = TRUE) %>% purrr::walk(knitr::read_chunk)
dir("package-demos", full.names = TRUE) %>% purrr::walk(knitr::read_chunk)
```

Check available chunks

```{r, eval = FALSE}
knitr:::knit_code$get()  # use this to check chunks in the current session
```


# Purpose

This is a cheatsheet for R. A place to put useful code, for reminding of how to handle different situations that come up.

# Datasets

```{r datasets-section, child="datasets.Rmd", cache=FALSE, eval=FALSE}
```

# R-language design

```{r r-design, child="r-design.Rmd", cache=FALSE, eval=FALSE}
```

# Standard Libraries

```{r standard-library, child="standard-library/index.Rmd", cache=FALSE, eval=FALSE}
```

```{r "initiate polymorphic exclude", echo = FALSE}
include <- function(obj, elms) UseMethod("include")
exclude <- function(obj, elms) UseMethod("exclude")
```

# Helpers

```{r Helpers, child="helpers.Rmd", cache=FALSE, eval=FALSE}
```

# Object Types

```{r Object-types, child="object-types/index.Rmd", cache=FALSE, eval=FALSE}
```

# Package Usages

```{r, echo=FALSE, eval=FALSE}
# TODO:
# 1. add vignette for lubridate - or look for better alternatives
# 2. improve vignette for forcats
# 3. should I use tibble? is it faster? or is just printing that's better?
# 4. strangeness going on with purrr? can't compile
```

```{r Package-vignettes, child="package-vignettes/index.Rmd", eval = FALSE, cache=FALSE}
```

# Cheats

```{r cheats, child="cheats.Rmd", cache=FALSE, eval=FALSE}
```

```{r comparing-objects, child="comparing-objects.Rmd", cache=FALSE, eval=FALSE}
```

# Plotting

```{r plotting, child="plotting.Rmd", cache=FALSE, eval=FALSE}
```

# Statistics

```{r Statistics, child="statistics/index.Rmd", eval = FALSE, cache=FALSE}
```

# Best practices

```{r, echo=FALSE, eval=FALSE}
# TODO:
```

```{r best-practices, child="best-practices/index.Rmd", eval = FALSE, cache=FALSE}
```

# workflows

```{r, echo=FALSE, eval=FALSE}
# TODO:
# 1. messed up indentation etc.
# 2. move section? perhaps compbine with best practices?
```

```{r workflows, child="workflows.Rmd", eval = FALSE, cache=FALSE}
```

# installation of packages

```{r installations, child="installations.Rmd", eval = FALSE, cache=FALSE}
```

# References
