## Object system

### Overview

R has a S3 and S4 Object system. S3 is preferred using polymorphism.
Ref: [adv-R](http://adv-r.had.co.nz/OO-essentials.html)

### Base types

R "base types" are like a C struct, has 1) object contents and 2) a type. This is not an object system as you can't add objects to it.

```{r, eval=FALSE}

f <- function() NULL
typeof(f) # [1] "closure"
is.function(f)

typeof(sum) # [1] "builtin" # for primitive functions
is.primitive(sum)

```

use `pryr::otype` to understand what kind of system an object follows

```{r, eval=FALSE}

pryr::otype(character()) # [1] "base"
pryr::otype(integer()) # [1] "base"
pryr::otype(numeric()) # [1] "base"

pryr::otype(matrix()) # [1] "base"
pryr::otype(list()) # [1] "base"
pryr::otype(c()) # [1] "base"

pryr::otype(factor()) # [1] "S3"
pryr::otype(ts()) # [1] "S3"
pryr::otype(data.frame()) # [1] "S3"
as.Date("1999-01-01") %>% pryr::otype() # [1] "S3"

df <- data.frame(x = 1:5, y = LETTERS[1:5])
pryr::otype(df)
pryr::otype(df$x)
pryr::otype(df$y) # factors are S3

pryr::otype(mean)

```

use `pryr::ftype`

```{r}

pryr::ftype(`%in%`)
pryr::ftype(sum)
pryr::ftype(t.data.frame)
pryr::ftype(t.test) # Tricky!
pryr::ftype(writeLines)
pryr::ftype(unlist)

pryr::ftype(mean)
methods(mean)
pryr::ftype(mean.Date)

```


### S3

#### methods

- query all methods for a S3 generic function
- query all methods S3 function available for a class

```{r, ""}

pryr::ftype(mean)
methods(mean)
methods(class = "Date")

```


#### generic function "UseMethod"

TODO: add example where you pass more arguments

base for `R`'s S3 object system using polymorphism, switching function (function / method dispatch) on the class of the called object.

```{r, "UseMethod"}

asEmpty <- function(object) UseMethod("asEmpty")
asEmpty.integer <- function(object) integer(0)
asEmpty.character <- function(object) character(0)

methods(asEmpty)

1:3 %>% asEmpty
letters %>% asEmpty

```

#### constructor function

example. think the factor and `data.frame` functions

```{r}

foo <- function(x) {
  if (!is.numeric(x)) stop("X must be numeric")
  structure(list(x), class = "foo")
  }

```

#### Use-case

adding custom classes to objects and using polymorphism

```{r}

mkNameFeatures <- function(obj) UseMethod("mkNameFeatures")
mkNameFeatures.se <- function(names) data.frame(name = names, namesCount = names %>% strsplit(" ") %>% purrr::map_dbl(length))
mkNameFeatures.ja <- function(names) data.frame(name = names, hasSpace = names %>% grepl(pattern = " ") )

initClass <- function(obj, cls) obj %>% structure(class = cls %>% append(class(obj)) %>% unique)

jaNames <- c("田中角栄", "杉本 レイ", "上杉謙信") %>% initClass("ja")
seNames <- c("Emil Erik Alexis Rehnberg", "stella", "balle AF raggare") %>% initClass("se")

jaNames %>% mkNameFeatures
seNames %>% mkNameFeatures

jaNames %>% mkNameFeatures %>% initClass("ja")

```

