
### Overview

- `ggplot2` will have precedence to base standard library `graphics::plot`

### helpers

`cut_width` is a helper function for easier cutting of numerical values.

```{r ggplot_cut_width}
library(magrittr)
set.seed(12)
(rr <- runif(10, max = 9) %>% round)
rr %>% ggplot2::cut_width(3)
```

### ggplot in functions

with `ggplot` in a function, if you're passing arguments for `aes` use `aes_string`.

```{r}
library(ggplot2)
cars <- within(mtcars, { cyl <- as.factor(cyl) })
renderBoxplot <- function(yvar){
  ggplot(cars, aes_string(x = "cyl", y = yvar)) + geom_boxplot()
}
renderBoxplot("mpg")
renderBoxplot("hp")
```

### Typical

```{r}
library(ggplot2)
(ggplot(airquality,
        aes(x = paste(Month, Day, sep="-"), y = Wind))
  + geom_point(size = .7)
  + geom_line(size = .5, colour = "blue")
  + ylab("Wind (mph)")
  + xlab("Days")
  + ylim(0, 25)   # sets limit on y-axis
  + ggtitle("Wind over time")
  + geom_hline(aes(yintercept = median(Wind)), colour = "red")
  + scale_x_discrete(breaks=NULL)
)
```

### line plot

line plots for e.g. a grouped (group by) data doesn't work since there's only one group, but ggplot expects there to be more than one group and it draws lines within the group. Hence, create a dummy group for ggplot.

typical error: `geom_path: Each group consist of only one observation. Do you need to adjust the group aesthetic?`

```{r}
library(ggplot2)
library(magrittr)
library(dplyr)
( airquality
  %>% group_by(Month)
  %>% summarise(wind = mean(Wind))
  %>% ggplot(aes(y = wind, x = Month, group = 1))
    + geom_point()
    + geom_line()
 )
```

with line groups

```{r}
library(ggplot2)
( ggplot( within(airquality, {Month %<>% as.factor})
         ,aes( x = Day
              ,y = Temp
              ,group = Month
              ,colour = Month)
       )
  + geom_line(size = .5)
  + geom_point(size = .7, fill = "white")
)
```

### non-numeric argument

typical error: `non-numeric argument to binary operator`

this can be due to many reasons! most common is some syntax error, check trailing `+` and paratheses.

### Histogram

```{r}
ggplot(airquality, aes(Wind)) + geom_histogram()
```

### Boxplot

```{r}
library(ggplot2)
(ggplot(mpg, aes(class, hwy))
 + geom_boxplot(outlier.size = 0.5)
)
```

to deal with horizontal labels being to many, you can angle (diagonal, twist, turn, transpose, flip) the labels

```{r}
aq <-
  within(airquality, {
           mthStg = Day %>% cut(breaks = c(0, 10, 20,32), labels = c("A-ge","B-chuu","C-jou"))
           yrStg = paste(Month, mthStg, sep = "-")
  })
( ggplot(aq, aes(yrStg, Temp))
  + geom_boxplot(outlier.size = .5)
  + ylab("Temperature")
  + xlab("Year Stage")
  + theme(axis.text.x = element_text(angle = 90, hjust = 1))
  )
```

### Dendrogram

See hierarchical clustering for base `plot` and `hclust`

```{r "dendrogram_ggplot"}

library(ggplot2)
library(ggdendro)
arrestsHc <- USArrests %>% dist %>% hclust("ave")
arrestsHc %>% ggdendrogram
arrestsHc %>% ggdendrogram(rotate = TRUE)

```

```{r "hc dendrogram with graphics::plot"}

arrestsHc %>% plot
arrestsHc %>% rect.hclust(5)
arrestsHc %>% plot
arrestsHc %>% rect.hclust(h = 40, which = c(2, 6), border = 3:4) # border is for colors

arrestsHc %>% plot
arrestsHc %>% rect.hclust(3)
arrestsHc %>% rect.hclust(h = 30, which = c(3, 7), border = 3:4)

```

extracting dendrogram plot data

```{r "dendrogram data extraction"}

library(ggplot2)
arrestsDendroDt <- arrestsHc %>% as.dendrogram %>% dendro_data(type = "rectangle")
ggplot(segment(arrestsDendroDt)) +
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend)) +
  coord_flip() +
  scale_y_reverse(expand = c(.2, 0))
.Last.value + theme_dendro() # remove axises and background

ggplot(segment(arrestsHc %>% as.dendrogram %>% dendro_data(type = "triangle"))) +
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend)) +
  coord_flip() +
  scale_y_reverse(expand = c(.2, 0))

```

regression tree diagrams

```{r}
devtools::install_github("cran/tree")
```


```{r ""}

library(tree)
data(cpus, package = "MASS")
model <- tree(log10(perf) ~ syct + mmin + mmax + cach + chmin + chmax
              ,data = cpus)
tree_data <- dendro_data(model)
ggplot(segment(tree_data)) +
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend, size = n)
               ,colour = "blue", alpha = 0.5) +
  scale_size("n") +
  geom_text(data = label(tree_data)
            ,aes(x = x, y = y, label = label), vjust = -0.5, size = 3) +
  geom_text(data = leaf_label(tree_data)
            ,aes(x = x, y = y, label = label), vjust = 0.5, size = 2) +
  theme_dendro()

ggplot(segment(tree_data)) +
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend) ,colour = "blue") +
  scale_size("n") +
  geom_text(data = label(tree_data)
            ,aes(x = x, y = y, label = label), vjust = -0.5, size = 3) +
  geom_text(data = leaf_label(tree_data)
            ,aes(x = x, y = y, label = label), vjust = 0.5, size = 2) +
  theme_dendro()

ggplot() +
  geom_segment(data = tree_data$segments
               ,aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_text(data = tree_data$labels
            ,aes(x = x, y = y, label = label), size = 3, vjust = 0) +
  geom_text(data = tree_data$leaf_labels
            ,aes(x = x, y = y, label = label), size = 3, vjust = 1) +
  theme_dendro()

```

```{r "trees with rpart"}

data("kyphosis", package="rpart")
rpart_data <-
  rpart::rpart(Kyphosis ~ Age + Number + Start
               ,method = "class"
               ,data = kyphosis) %>% dendro_data

ggplot() +
  geom_segment(data = rpart_data$segments
               ,aes(x = x, y = y, xend = xend, yend = yend)
               ,colour = "blue") +
  geom_text(data = rpart_data$labels
            ,aes(x = x, y = y, label = label), size = 3, vjust = 0) +
  geom_text(data = rpart_data$leaf_labels
            ,aes(x = x, y = y, label = label), size = 3, vjust = 1) +
  theme_dendro()

co2_rpart_data <-
  rpart::rpart(Treatment ~ Type + conc + uptake
               ,method = "class"
               ,data = CO2) %>% dendro_data

ggplot() +
  geom_segment(data = co2_rpart_data$segments
               ,aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_text(data = co2_rpart_data$labels
            ,aes(x = x, y = y, label = label), size = 3, vjust = 0) +
  geom_text(data = co2_rpart_data$leaf_labels
            ,aes(x = x, y = y, label = label), size = 3, vjust = 1) +
  theme_dendro()

```

### Facet grid

- use the formula to tell how and what you want to stratify the data on.
- Plot is split differently if `Species` is on the x-side
- some added smoothing
- use y and x side if you want to split on multiple categorical variables

```{r}
( ggplot(iris, aes(y = Sepal.Length,
                   x = Sepal.Width))
  + facet_grid(Species ~ .)
  + geom_smooth()
  )
```

### Themes

```{r ggplot_themes}
library(dplyr)
library(ggplot2)
diamondsTrunc <- diamonds[1:5000,]

( ggplot(data=diamondsTrunc, aes(carat,price ))
  + geom_point(aes(colour= color))
  + theme_bw()
  )

( ggplot(data=diamondsTrunc, aes(carat,price ))
  + geom_point(aes(colour= color))
  + theme_classic()
  )

( ggplot(data=diamondsTrunc, aes(carat,price ))
  + geom_point(aes(colour= color))
  + theme_bw()
  )

( ggplot(data=diamondsTrunc, aes(carat,price ))
  + geom_point(aes(colour= color))
  + theme_dark()
  )

( ggplot(data=diamondsTrunc, aes(carat,price ))
  + geom_point(aes(colour= color))
  + theme_gray()
  )

( ggplot(data=diamondsTrunc, aes(carat,price ))
  + geom_point(aes(colour= color))
  + theme_light()
  )

( ggplot(data=diamondsTrunc, aes(carat,price ))
  + geom_point(aes(colour= color))
  + theme_linedraw()
  )

( ggplot(data=diamondsTrunc, aes(carat,price ))
  + geom_point(aes(colour= color))
  + theme_minimal()
  )
```

### Matrices

#### Overview

useful to see pairwise correlation of variables.

#### graphics::pairs

pairs with added loess smoother in lower and correlation in upper

```{r}
panel.cor <- function(x, y, digits=2, prefix="", cex.cor, ...)
{
    usr <- par("usr"); on.exit(par(usr))
    par(usr = c(0, 1, 0, 1))
    r <- abs(cor(x, y))
    txt <- format(c(r, 0.123456789), digits=digits)[1]
    txt <- paste(prefix, txt, sep="")
    if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
    text(0.5, 0.5, txt, cex = cex.cor * r)
}

pairs(~Sepal.Length+Sepal.Width+Petal.Length+Petal.Width, data=iris,
      lower.panel=panel.smooth, upper.panel=panel.cor,
      pch=20, main="Iris Scatterplot Matrix")
```

#### lattice

Has options to condition the scatterplot matrix on a factor.

```{r}
library(lattice)
super.sym <- trellis.par.get("superpose.symbol")
splom(~iris[1:4],
      groups = Species,
      data = iris,
      panel = panel.superpose,
      key = list(title = "Three Varieties of Iris",
                 columns = 3,
                 points = list(pch = super.sym$pch[1:3],
                 col = super.sym$col[1:3]),
                 text = list(c("Setosa", "Versicolor", "Virginica"))))
```

#### car

The `car` package can condition the scatterplot matrix on a factor, and optionally include lowess and linear best fit lines, and boxplot, densities, or histograms in the principal diagonal, as well as rug plots in the margins of the cells.

```{r, eval=FALSE, echo=FALSE}

dependencies <- c("minqa", "nloptr", "Rcpp", "RcppEigen", "lme4", "pbkrtest", "SparseM", "MatrixModels", "quantreg", "car")
devtools::install_github(paste("cran", dependencies, sep = "/"))

```

```{r}
car::scatterplotMatrix(~mpg+disp+drat+wt|cyl
                       ,data=mtcars
                       ,main="Three Cylinder Options")
```

#### gclus

The `gclus` package provides options to rearrange the variables so that those with higher correlations are closer to the principal diagonal. It can also color code the cells to reflect the size of the correlations.

```{r, eval=FALSE, echo=FALSE}
devtools::install_github("cran/gclus")
```

```{r}

dta <- mtcars[c(1,3,5,6)] # get data
dta.r <- abs(cor(dta)) # get correlations
dta.col <- gclus::dmat.color(dta.r) # get colors

```

reorder variables so those with highest correlation are closest to the diagonal

```{r}

dta.o <- gclus::order.single(dta.r)
gclus::cpairs(dta, dta.o, panel.colors=dta.col, gap=.5,
              main="Variables Ordered and Colored by Correlation" )

```

#### gpairs

```{r, eval = FALSE}

dependencies <- c("barcode", "lmtest", "vcd", "gpairs")
devtools::install_github(paste("cran", dependencies, sep = "/"))

```

```{r}
if (allexamples) {
  y <- data.frame(A=c(rep("red", 100), rep("blue", 100)),
                  B=c(rnorm(100),round(rnorm(100,5,1),1)),
                  C=runif(200),
                  D=c(rep("big", 150), rep("small", 50)),
                  E=rnorm(200))
  gpairs::gpairs(y)
}

if (allexamples) {
  data(iris)
  gpairs::gpairs(iris)
  gpairs::gpairs(iris, upper.pars = list(scatter = 'stats'),
         scatter.pars = list(pch = substr(as.character(iris$Species), 1, 1),
                             col = as.numeric(iris$Species)),
         stat.pars = list(verbose = FALSE))
  gpairs::gpairs(iris, lower.pars = list(scatter = 'corrgram'),
         upper.pars = list(conditional = 'boxplot', scatter = 'loess'),
         scatter.pars = list(pch = 20))
}

if (allexamples) {
  data(Leaves)
  gpairs::gpairs(Leaves[1:10], lower.pars = list(scatter = 'loess'))
  gpairs::gpairs(Leaves[1:10], upper.pars = list(scatter = 'stats'),
         lower.pars = list(scatter = 'corrgram'),
         stat.pars = list(verbose = FALSE), gap = 0)
  gpairs::corrgram(Leaves[,-33])
}
```

#### GGally

```{r, eval=FALSE, echo=FALSE}

dependencies <- c("prettyunits", "progress", "reshape", "GGally")
devtools::install_github(paste("cran", dependencies, sep = "/"))

```

not sure how well this package is working.
Examples given in `?GGally::ggpairs` doesn't seem to work very well (or at all).

```{r ggally_matrix_plot, eval = FALSE}
library(GGally)
if (allexamples){
  ds <- read.csv("for-gally.csv")
  ds$sex <- ifelse(ds$female==1, "female", "male") %>% as.factor
  ds$housing <- ifelse(ds$homeless==1, "homeless", "housed") %>% as.factor
  GGally::ggpairs(ds,
          columns=c("i1", "cesd", "housing", "sex"), # list the factor variables first to get better boxplots
          diag=list(continuous="density",
                    discrete="bar"),
          axisLabels="show")
}

if (allexamples){
  data(tips, package = "reshape")
  GGally::ggpairs(tips[, 1:3])
  GGally::ggpairs(tips, 1:3, columnLabels = c("Total Bill", "Tip", "Sex"))
  GGally::ggpairs(tips, 1:3, upper = "blank")
  GGally::ggpairs(tips[, 1:3], axisLabels="internal") # Only Variable Labels on the diagonal (no axis labels)
  GGally::ggpairs(tips[, 1:3], axisLabels="none") # Only Variable Labels on the outside (no axis labels)

  GGally::ggpairs(
    tips[, c(1, 3, 4, 2)],
    upper = list(continuous = "density", combo = "box"),
    lower = list(continuous = "points", combo = "dot")
  )
}

if (allexamples){
  data(diamonds, package="ggplot2")
  diamonds.samp <- diamonds %>% sample_n(30)

  GGally::ggpairs(
    diamonds.samp,
    columns = "carat cut clarity depth" %>% wordArray,
    mapping = ggplot2::aes(color = cut),
    upper = list(continuous = wrap("density", alpha = 0.5),
                 combo = "box"),
    lower = list(continuous = wrap("points", alpha = 0.3),
                 combo = wrap("dot", alpha = 0.4)),
    title = "Diamonds"
  )
}

custom_car <- GGally::ggpairs(mtcars[, c("mpg", "wt", "cyl")], upper = "blank", title = "Custom Example")

```

ggplot example taken from example(geom_text)

```{r eval = FALSE}
plot <- ggplot2::ggplot(mtcars, ggplot2::aes(x=wt, y=mpg, label=rownames(mtcars)))
plot <- plot +
  ggplot2::geom_text(ggplot2::aes(colour=factor(cyl)), size = 3) +
  ggplot2::scale_colour_discrete(l=40)
custom_car[1, 2] <- plot
personal_plot <- GGally::ggally_text(
  "ggpairs allows you\nto put in your\nown plot.\nLike that one.\n <---"
)
custom_car[1, 3] <- personal_plot
custom_car
```

#### ggplot2::plotmatrix

Depricated.

```{r, eval = FALSE}
library(ggplot2)
plotmatrix(with(iris, data.frame(Sepal.Length, Sepal.Width, Petal.Length, Petal.Width)))
```

### ggplot2 with multiple datasets and colors

```{r}
library(dplyr)
library(magrittr)
library(ggplot2)
dpassed <- rep(c("Yes", "No"), each=1000)
dspeed <- c(rnorm(1000, 2),
           rnorm(1000, 1))
lvlData <- data.frame(passed = dpassed, speed = dspeed)
samples <- factor(c("Passed", "Total"))

```

TODO: FIX ME

```{r , eval = FALSE}
(
  ggplot(lvlData %>% filter(passed == "Yes"),
         aes(x=speed, color="Sample", shape="Sample", linetype="Sample"))
  + stat_ecdf(data=lvlData,
              aes(x=speed, color="Total", shape="Total", linetype="Total"))
  + stat_ecdf(aes(color="Passed", shape="Passed", linetype="Passed"))
  + xlab("")
  + ylab("inclusion")
  + labs(title="Average days to completion")
  + scale_color_manual(breaks=samples, values=c("green", "red"))
  + scale_shape_manual(breaks=samples, values=c(16, 16))
  + scale_linetype_manual(breaks=samples, values=c(1, 1))
  + labs(color = "dataset")
  + theme(legend.title=element_blank())
)
```

### add linear regression line

```{r}
dlmDtExample <-
  structure(list(date =
                 structure(c(3407, 3438, 3468, 3499, 3530, 3560, 3591, 3621, 3652, 3683, 3712, 3743, 3773, 3804, 3834,
                             3865, 3896, 3926, 3957, 3987, 4018, 4049, 4077, 4108, 4138, 4169, 4199, 4230, 4261, 4291,
                             4322, 4352, 4383, 4414, 4442, 4473, 4503, 4534, 4564, 4595, 4626, 4656, 4687, 4717, 4748,
                             4779, 4807, 4838, 4868, 4899, 4929, 4960, 4991, 5021, 5052, 5082, 5113, 5144, 5173, 5204,
                             5234, 5265, 5295, 5326, 5357), class = "Date"),
                 gas = c(4.9, 4.5, 3.1, 3.1, 4.6, 4.8, 6.7, 8.6, 8.3, 7.2, 9.2, 5.5, 4.7, 4.4,
                         3.4, 2.8, 4, 5.1, 6.5, 9.2, 7.7, 7.7, 8.9, 5.7, 5, 4.5, 3.3,
                         2.8, 4, 5.6, 6.6, 10.3, 8.5, 7.9, 8.9, 5.4, 4.4, 4, 3, 3.1, 4.4,
                         5.5, 6.5, 10.1, 7.7, 9, 9, 6.5, 5.1, 4.3, 2.7, 2.8, 4.6, 5.5,
                         6.9, 9.5, 8.8, 8.7, 10.1, 6.1, 5, 4.5, 3.1, 2.9, 4.8)),
            .Names = c("date", "gas"),
            row.names = c(NA, -65L),
            class = "data.frame")

( ggplot(dlmDtExample, aes(x = date, y = gas))
  + geom_line()
  + ggtitle("UK gas consumption: amount in 106 tonnes coal equivalent")
  + geom_smooth(method = "lm", se = FALSE)
)

```

or a polynomial

```{r}
( ggplot(dlmDtExample, aes(x = date, y = gas))
  + geom_line()
  + ggtitle("UK gas consumption: amount in 106 tonnes coal equivalent")
  + geom_smooth(method = "lm", se = FALSE, formula = y ~ poly(x, 4))
)

```

