## Overview

Section for R-language design and concepts. Packages, code, data loading

## loading / unloading packages

in order to attach / load / require packages use `library` (or `require` if you like problems).

to detach / unload / remove a package from a session you can use `detach`

```{r}
exists("is_in") # from magrittr package
detach(package:magrittr)
exists("is_in")
library(magrittr)
exists("is_in")
```

## error handling

use `tryCatch`

with error handler

```{r "error / exception handling", echo = TRUE}

library(magrittr)
myError <- simpleError("エッラー！")

tryCatch({
  print(1);
  stop(myError)
  print(2)
}, error = function(err){
  message("error function messages:")
  (err %>% print)
}, finally = {
  print("always finish run with finally()")
  message("finished tryCatch")
})

```

with warning handler

```{r}

tryCatch({
  print(1)
  warning("ACHTUNG! ACHTUNG!")
  print(2)
}, warning = function(war){
  message("warning function messages:")
  (war %>% print)
}, finally = {
  print("always finish run with finally()")
  message("finished tryCatch")
})

```

`stop` returns exit code 1 (check with `$?` in ruby and `$?.exitstatus` especially)

you can also use the `tools` package. but `pskill` does not returns a 0 exit status. `pskill` seems to be rather rough around the edges, use with care.

```{r}
# tools::pskill(Sys.getpid(), tools::SIGINT)
```

## Control Flow

E.g. if / else statements

```{r "control flow, all the below can be parsed"}

if(1 == 2)
  print("1 == 2") else
if(2 == 2)
  print("2 == 2") else
  print("ALL FALSE")

if(1 == 2) {
  print("1 == 2")
} else if(2 == 2) {
    print("2 == 2")
  } else {
    print("ALL FALSE")
  }

if(1 == 2) {
  print("1 == 2")
} else {
  if(2 == 2) {
    print("2 == 2")
  } else {
    print("ALL FALSE")
  }
}

```

## ellipsis

`...` / ellipsis or "dot-dot-dot"

use an arbitrary number of arguments to a function

```{r}
a <- c(1,  2,  NA, 4, NA)
b <- c(NA, NA, NA, 5, 6)
c <- c(7,  8,  NA, 9, 10)
```

catch the arguments in a list

```{r}
elTrail1 <- function(...) list(...)
```
catch the arguments in a data frame

```{r}
elTrail2 <- function(...) data.frame(...)
```

extract the first argument

```{r}
elTrail3 <- function(...) ..1
```

extract the second argument

```{r}
elTrail4 <- function(...) ..2
```

## Environments

```{r}
exists("abc")
abc <- 2
exists("abc")
exists("pi", envir = emptyenv())
exists("glm", envir = as.environment("package:base"))
exists("glm", envir = as.environment("package:stats"))
```

## dump object to executable R code

`dput` is your friend. NOTE: not suitable for S4 objects. try `writeRDS` or `save` for that.

there's some gotchas though. When you're dumping a factor - the output might end up crazy long. Use `as.character` to get what you're looking for.

```{r "dput gotcha"}
library(dplyr)
(USArrests %<>% mutate(state = USArrests %>% row.names %>% as.factor)) %>% str
USArrests$state %>% head(2) %>% dput
USArrests$state %>% head(2) %>% as.character %>% dput
```

## parse CL args

parsing command line arguments / parameters

- use `#!/usr/bin/env Rscript` on top of the executable file (`tca.r` let's say)
- call `./tca.r --args a=1 name=\'Billy\'`
- use the `commandArgs`, `parse` and `eval` functions to read the parameters into your script
- sometimes if `trailingOnly` is set, `--args` is still in there

```{r "commandArgs"}
library(magrittr)
arguments <- c("a=1", "name=\'Billy\'") # commandArgs(trailingOnly = TRUE)
for(argument in arguments){
  eval(parse(text = argument))
}
```

into a `data.frame`. This works but then all columns are strings

```{r "command line arguments into data.frame"}
arguments %>% strsplit("=") %>% sapply(function(strings) purrr::set_names(strings[2], strings[1]) )
```

use `jsonlite` instead by sending in JSON data. E.g. executing:
`./predict-co2uptake '[{"Type": "Quebec", "Treatment": "nonchilled", "conc": 350}]'`
on the command line

```{r "command line arguments into data.frame using jsonlite"}
jsonlite::fromJSON('[{"Type": "Quebec", "Treatment": "nonchilled", "conc": 350}]') %>% str
```

```{r object-system, child="object-system.Rmd"}
```
