## devtools

### Overview

Section for devtools basics and qwirks. Use it for creating packages, downloading them etc.

### install.github

install packages from github

```{r "devtools::install_github", eval = FALSE}
devtools::install_github("cran/purrr")
devtools::install_github("hadley/stringr")
devtools::install_github("rstudio/rmarkdown")
devtools::install_github("rstudio/rmarkdown", ref = '00b58f6') # v1.5
devtools::install_github(c("cran/cli", "cran/utf8", "cran/pillar", "cran/tibble"))
devtools::install_github("rstudio/rmarkdown", ref = '79a741b') # v1.1
```

### create

generates a R-package template

```{r "devtools::create", eval = FALSE}
devtools::create("path/to/package/pkgname")
devtools::create("~/code/R/packages/niftytools")
```

### add imports

adds `magrittr` dependency to import field to DESCRIPTION file in a package

```{r "devtools::use_package", eval = FALSE}
devtools::use_package("magrittr")
```

### document

writes roxygen2 documentation, writing NAMESPACE and .Rd (documentation) files.

```{r "devtools::document", eval = FALSE}
devtools::load_all()
devtools::document()
```

### load package contents

Use `devtools::load_all()` function to load the package during development.

### using testthat

Run `devtools::use_testthat()` to set-up testing during package development.

### run tests

Run `devtools::test()` to run the package tests.

