## dplyr

### Overview

Section for dplyr basics and qwirks

### programming

evaluating variable names and values. This is for dplyr 6.0+

```{r "dplyr programming"}

dta <- data.frame(
   v1 = sample(5,10, replace = TRUE)
  ,v2 = sample(5,10, replace = TRUE)
)

```

dplyr filter is a little bit qwirkier.

Note the parentheses around the quo eval, it's required in this situation.

```{r}

myVar <- rlang::quo(v1)
dta %>% dplyr::filter((!!myVar) == 1)

```

while not required in this situation

```{r}

myVal <- 1
dta %>% dplyr::filter(v1 == !!myVal)

```

not that col is a string / character

```{r}

which_col <- "v1"
which_val <- 1

dta %>% dplyr::filter((!!rlang::sym(which_col))==which_val)

```

OR

```{r}

dta %>% dplyr::filter(UQ(rlang::sym(which_col))==which_val)

```

old syntax still works

```{r}

dta %>% dplyr::filter_(.dots= paste0(which_col, "== ", which_val))

```

### subsetting rows

observations

```{r "dplyr filter"}
dplyr::filter(iris, Sepal.Length > 7)
```

remove duplicate rows

```{r "dplyr distinct"}
iris %>% str
dplyr::distinct(iris) %>% str
```

Randomly select fraction of rows.

```{r "dplyr sample_frac"}
dplyr::sample_frac(iris, 0.05, replace = TRUE)
```

Randomly select n rows.

```{r "dplyr sample_n"}
dplyr::sample_n(iris, 10, replace = TRUE)
```

Select rows by position.

```{r "dplyr slice"}
dplyr::slice(iris, 10:15)
```

Select top n rows based on some column

```{r "dplyr top_n"}
CO2 %>% top_n(2, uptake)  # select top 2 observations with highest uptake
CO2 %>% top_n(2, conc)    # conc has a lot of ties
CO2 %>% top_n(2, -uptake) # select top 2 observations with LOWest uptake
```

### selects

- different ways of using `dplyr::select`
- use array with `one_of` to select columns from data frame

```{r}
library(magrittr); library(dplyr)
airquality %>% head %>% select(Temp, Wind)
airquality %>% head %>% select_("Temp", "Wind")
var1 <- "Temp"; var2 <- "Wind"
airquality %>% head %>% select_(var1, var2)

wordArray <- function(string){ string %>% strsplit("[[:space:]]+") %>% unlist }
columns <- "Temp Wind" %>% wordArray
airquality %>% head %>% select(one_of(columns))
# airquality %>% head %>% select_(one_of(columns)) # ERRORS: can not find function `one_of`
```

### filter

If you get weird time-series results then standard library `filter` is most likely first in the namespace.

Use `%in%` or `is_in` for filtering on multiple items.

```{r}
library(dplyr)
set.seed(12)
threshold <- 0.7
dts <- data.frame(a1 = runif(20) %>% round(1), b1 = runif(20) %>% round(1))
dts %<>% mutate(b1Lvl = b1 %>% cut(breaks =c(0, threshold, 1), include.lowest = TRUE, labels = c("low", "high")) )

dts %>% filter(a1 %>% is_in(c(0, 0.2)))
dts %>% filter(a1 %in% c(0, 0.2))

dts %>% filter_("b1 > .7")
dts %>% filter_(paste("b1 >", threshold))
dts %>% filter_("b1Lvl == 'high'")
asQuoted <- function(arg) paste0("'", arg, "'")
b1Level <- "high"
dts %>% filter_(paste("b1Lvl == ", b1Level %>% asQuoted))
```

