## purrr

### Overview

Overview of purrr package.

### use-case

- `map()` makes a list.
- `map_lgl()` makes a logical vector.
- `map_int()` makes an integer vector.
- `map_dbl()` makes a double vector.
- `map_chr()` makes a character vector.

```{r "purrr use case"}
library(purrr)

mtcars %>%
  split(.$cyl) %>% # from base R
  map(~ lm(mpg ~ wt, data = .)) %>%
  map(summary) %>%
  map_dbl("r.squared")
```

```{r "two level split, 2nd split only on half"}
dff <-
  data.frame(F1 = gl(2,3, labels = c("a", "b"))
             ,F2 = gl(3,2, labels = LETTERS[1:3])
             ,r1 = rnorm(6)
             )
stratDff <- dff %>% split(.$F1)
stratDff$b %<>% split(.$F2)

```

### map

```{r "purrr::map"}
library(purrr)

sums <-
  mtcars %>%
    split(.$cyl) %>% # from base R
    map(~ lm(mpg ~ wt, data = .)) %>%
    map(summary)

sums %>% map_dbl("r.squared")
sums %>% map_chr("r.squared")
```

to add names use pipe `set_names` after

```{r "purrr::map with names"}
sums %>% map("call") %>% purrr::set_names(paste0("CylinderCnt_", c(4,6,8)))
```

`.id` argument is available for `purrr::map_df`

```{r ".id argument only works with map_df"}

purrr::rerun(5, rnorm(100)) %>%
  set_names(paste0("sim", 1:5)) %>%
  purrr::map_df(~ dplyr::tibble(value = .x, step = 1:100), .id = "simulation")

```

### imap - map with index / names

iterate with names for `map`, `walk` etc.
`imap(lst, .f)` is short for `map(lst, names(list), .f)`

```{r}

c(mohammad = "^moh$|^muhamed$|^mohamed$",ahmad = "^ahmed$|^ahmat$") %>% purrr::imap(~ paste0("regex from ", .y, ": ", .x))

c(mohammad = "^moh$|^muhamed$|^mohamed$",ahmad = "^ahmed$|^ahmat$") %>% purrr::imap_chr(~ paste0("regex from ", .y, ": ", .x))

```



### partial

Curry a function. i.e. prefill arguments for a function. by default the prefilled arguments are evaluated lazily.

```{r "purrr::partial"}
compact1 <- function(x) discard(x, is.null)
compact2 <- partial(discard, .p = is.null)
```

and the generated source code is very similar to what we made by hand

```{r}
compact1
compact2

f <- partial(runif, n = rpois(1, 5))
f
f()
f()
```

You can override this by saying .lazy = FALSE

```{r}
f <- partial(runif, n = rpois(1, 5), .lazy = FALSE)
f
f()
f()
```

### prepend

companion to `append`

```{r purrr::prepend}
x <- as.list(1:3)
x %>% append("a")
x %>% prepend("a")
```

### generate random discrete numbers

```{r "purrr::rdunif"}
rdunif(1e3, 10) %>% table
```

### reduce

similar to `base::Reduce`. except the argument order makes sense :D

```{r "purrr::reduce"}
1:10 %>% reduce(`*`)
```

### rerun

rerun some code and `n` times and output the result in a list of length `n`

```{r "purrr::rerun"}

3 %>% rerun(rnorm(5))
2 %>% rerun(nrm = rnorm(3), unif = runif(3))

```

### dealing with failure

use `safely` function.
usage is very restrictive!

```{r "purrr::safely"}
safe_log <- safely(log)
safe_log(10)
safe_log("a")

loggedList <- list(1, 10, "a") %>% map(safely(log))
loggedList %>% str

loggedList %>% transpose %>% str
```

### scalar predicates

predicates for an object being of a type and being scalar (i.e. length 1)

- `is_scalar_list(x)`
- `is_scalar_atomic(x)`
- `is_scalar_vector(x)`
- `is_scalar_numeric(x)`
- `is_scalar_integer(x)`
- `is_scalar_double(x)`
- `is_scalar_character(x)`
- `is_scalar_logical(x)`

### setting names

snake case wrapper for `setNames`, tweaked defaults and stricter argument checks

```{r "purrr::set_names"}
set_names(1:4, letters[1:4])
set_names(letters[1:4]) # names is defaulting to the vector itself
```

### splice

function for appending to a list, or `splicing` all arguments into a list.

probably `update_list` is preferred though due to the edge-case shown below

```{r}
library(purrr)
```

want to an arg3 to inputs

```{r}
inputs <- list(arg1 = "a", arg2 = "b")
```

below is not sufficient

```{r}
list(inputs, arg3 = c("c1", "c2")) %>% str()
c(inputs, arg3 = c("c1", "c2")) %>% str()
```

use `splice`. `splice()` concatenates the elements of inputs with arg3

```{r}
splice(inputs, arg3 = c("c1", "c2")) %>% str()
```

edge-case when two elements get named the same

```{r}
splice(inputs, arg2 = c("c1", "c2")) %>% str()
```

use update_list instead

```{r}
update_list(inputs, arg2 = c("c1", "c2")) %>% str()
```

you can append but it's clunky

```{r}
append(inputs, list(arg3 = c("c1", "c2")))
```

or worse with a complete statement, mutating inputs

```{r}
inputs[["arg3"]] <- c("c1", "c2")
inputs
```

### split/order/sort by list components

DEPRECATED see [github release](https://github.com/tidyverse/purrr/tree/v0.2.3)

```{r "purrr::split_by", eval = FALSE}

(l1 <- transpose(list(x = sample(4), y = letters[1:4])))
l1 %>% order_by("x") # strings can't be used anymore
l1 %>% sort_by("x")

(l2 <- rerun(5, g = sample(2, 1), y = rdunif(5, 10)))
l2 %>% split_by("g") %>% str()
l2 %>% split_by("g") %>% map(. %>% map("y"))

```

### transpose

transpose a list to change order of indexing of a symmetric list

```{r "purrr::transpose"}
safe_log <- safely(log)
loggedList <- list(1, 10, "a") %>% map(safe_log)
loggedList %>% str
loggedList %>% transpose %>% str
```

### type predicates

- `is_list(x)`
- `is_atomic(x)`
- `is_vector(x)`
- `is_numeric(x)`
- `is_integer(x)`
- `is_double(x)`
- `is_character(x)`
- `is_logical(x)`
- `is_null(x)`
- `is_function(x)`

### update a list

modification of a list. similar to `splice` but can overwrite elements with same name.

```{r "purrr::update_list"}

inputs <- list(arg1 = "a", arg2 = "b")
purrr::update_list(inputs, arg2 = c("c1", "c2")) %>% str()

```

instead of using splice

```{r}
splice(inputs, arg2 = c("c1", "c2")) %>% str()

purrr::update_list(inputs, arg3 = c("c1", "c2")) %>% str()

SpecialArguments <- list(arg1 = c("c1", "c2"))
```

make it a deeper list

```{r}
purrr::update_list(inputs, SpecialArguments = SpecialArguments) %>% str
```

update list contents with overwrites using splicing !!! or `UQS()`

```{r}
purrr::update_list(inputs, !!! SpecialArguments) %>% str
```

if you add a list to list though use good old append

```{r}
spArgList <- list(specialArgs = c("c2", "c3"))
```

below broke

```{r, eval = FALSE}
inputs %>% purrr::update_list(spArgList) # will spit out inputs
```

```{r}
inputs %>% append(spArgList)
```

### when

pattern match with conditions using `when`. Matching the 2nd condition.

```{r "purrr::when"}
1:10 %>%
  when( sum(.) <=  50 ~ sum(.)
       ,sum(.) <= 100 ~ sum(.)/2
       ,~ 0
       )
```

matching both the first and 2nd conditions > first condition wins!

```{r}
1:10 %>%
  when( sum(.) <=   x ~ sum(.)
       ,sum(.) <= 2*x ~ sum(.)/2
       ,~ 0
       ,x = 60
       )
```

doesn't match the condition so run default

```{r}
iris %>%
  subset(Sepal.Length > 10) %>%
  when( nrow(.) > 0 ~ .
       ,~ iris %>% head(10)
       )
```

### iterating over multiple vectors simultaneously

use `pmap` to iterate over an arbitrary number of arguments / vectors at the same time.

```{r "purrr::pmap"}
library(purrr)
mu  <- list(5, 10, -3)
sgm <- list(1, 5, 10)
n   <- list(1, 3, 5)
```

with two arguments, you can use `map2`

```{r}
map2(mu, sgm, rnorm, n = 5) %>% str()
```

for more arguments, use a list and `pmap`.

no names means order is significant!!

```{r}
argsList <- list(n, mu, sgm)
argsList %>% pmap(rnorm)
```

use a named list for more comprehensable code

```{r}
argsList <- list(mean = mu, sd = sgm, n = n)
argsList %>% pmap(rnorm)

mu  <- list(5, 10, -3)
sigma <- list(1, 5, 10)
n <- list(1, 3, 5)
args1 <- list(n, mu, sigma)
args1 %>%
  pmap(rnorm) %>%
  str()
```

### walk

use `walk`/`pwalk` to iterate but you're interested in the side-effect instead of the returning value.

e.g. you have a list of plot objects and a list of filenames you want the plots saved to

```{r "purrr:walk"}
library(ggplot2)
plots <- mtcars %>%
  split(.$cyl) %>%
  map(~ggplot(., aes(mpg, wt)) + geom_point())
paths <- stringr::str_c(names(plots), ".pdf")

pwalk(list(paths, plots), ggsave, path = tempdir())
```

### get-attr

read attributes infix

```{r "purrr::get-attr"}
factor(1:3) %@% "levels"
mtcars %@% "class"
```

### empty?

`is_empty` function for vectors or lists

```{r purrr::is_empty}
is_empty(NULL)
is_empty(list())
is_empty(list(NULL))
```

### keep / discard / compact

usually called `filter`/`select`/`find_all` and `reject`/`drop` in other languages. (these names are taken however)

`compact` works well with both lists and vectors

```{r "purrr::keep"}
rep(10, 10) %>%
  map(sample, 5) %>%
  keep(function(x) mean(x) > 6)
```

Or use a formula

```{r}
rep(10, 10) %>%
  map(sample, 5) %>%
  keep(~ mean(.x) > 6)
```

Using a string instead of a function will select all list elements.
Where that subelement is `TRUE.`

```{r}
(x <- rerun(5, a = rbernoulli(1), b = sample(10)))
x %>% keep("a")
x %>% discard("a")
```

filter list on is.integer

```{r}
list(a = 1:3
     ,b = gl(4,3)
     ,d = integer()
     ,e = letters
     ) %>% purrr::keep(is.integer)
```

compact keeps all non-NULL and empty items

```{r}
list( a = 1
     ,b = NULL
     ,c = FALSE
     ,d = "FALSE"
     ,e = integer()
     ,f = character()
     ,g = list(NULL)
     ,h = NA
     ) %>% (purrr::compact)

c(1, NULL, 3, NA, 5) %>% compact
c(1, NULL, 3, NA, 5) %>% purrr::compact()
```

### conditional map

use `map_if` with predicate function and `map_at` with names or index

```{r "purrr::map_if"}
iris %>%
  purrr::map_if(is.factor, as.character) %>%
  str()
```

Specify which columns to map with a numeric vector of positions:

```{r}
mtcars %>% map_at(c(1, 4, 5), as.character) %>% str()
```

Or with a vector of names:

```{r}
mtcars %>% map_at(c("cyl", "am"), as.character) %>% str()
```

### lmap

apply a function to list-elements of a list, and must return a list. the length of the list going in can be of a different size to that one going out.

Let's write a function that returns a larger list or an empty list
depending on some condition. This function also uses the names
metadata available in the attributes of the list-element.

```{r "purrr::lmap"}
library(purrr)
maybe_rep <- function(e) {
  n <- rpois(1, 2)
  out <- rep_len(e, n)
  if (length(out) > 0) {
    names(out) <- paste0(names(e), seq_len(n))
  }
  out
}

x <- list(a = 1:4, b = letters[5:7], c = 8:9, d = letters[10])
```

The output size varies each time we map f()

```{r}
x %>% lmap(maybe_rep)
```

We can apply f() on a selected subset of x

```{r}
x %>% lmap_at(c("a", "d"), maybe_rep)
```

Or only where a condition is satisfied

```{r}
x %>% lmap_if(is.character, maybe_rep)
```

each element in the list is sent in as a list

```{r}
list(a = 1:2, b = letters[1:3]) %>% purrr::lmap(function(eal) { class(eal) %>% print; eal}) %>% invisible
```

making it possible to access the names of the list entries

```{r}
x %>% purrr::lmap(function(e) { names(e) %<>% paste0("_3"); e})
```

and you'll have to dig deeper to e.g. alter the names of the original list entries

```
list(a = c(B = 2, V = 3)) %>% lmap(function(eal) {names(eal[[1]]) %<>% paste0(0); eal})
```

### at depth

apply a function at a set depth of a nested list

`x %>% at_depth(0, fun)` is equivalent to `fun(x)`.

`x %>% at_depth(1, fun)` is equivalent to `map(x, fun)`.

`x %>% at_depth(2, fun)` is equivalent to `map(x, . %>% map(fun))`.

```{r "purrr::modify_depth prep"}

l1 <- list(
  obj1 = list(
    prop1 = list(param1 = 1:2, param2 = 3:4),
    prop2 = list(param1 = 5:6, param2 = 7:8)
    ),
  obj2 = list(
    prop1 = list(param1 = 9:10, param2 = 11:12),
    prop2 = list(param1 = 13:14, param2 = 15:16)
    )
  )

```

In the above list, "obj" is level 1, "prop" is level 2 and "param"
is level 3. To apply `sum()` on all params, we map it at depth 3:

```{r "purrr::at_depth DEPRECATED", eval = FALSE}

l1 %>% at_depth(3, sum)
```

map() lets us pluck the elements prop1/param2 in obj1 and obj2:

```{r}
l1 %>% map(c("prop1", "param2")) %>% str()
```

But what if we want to pluck all param2 elements? Then we need to act at a lower level:

```{r}
l1 %>% at_depth(2, "param2") %>% str()

```

```{r "purrr::modify_depth example"}

l1 %>% purrr::modify_depth(3, sum )
l1 %>% purrr::map(c("prop1", "param2"))
l1 %>% purrr::modify_depth(2, "param2")
l1 %>% purrr::modify_depth(2, ~ log(.$param2) )

```


### has_element

predicate function for checking inclussion / containing of objects in another.

works with lists and vectors. but there's some gotchas.

```{r "purrr::has_element"}

x <- list(1:10, 5, 9.9)
x %>% has_element(1:10)
x %>% has_element(3)
list(1, 2, 3) %>% has_element(2)
c(1, 2, 3) %>% has_element(2)
```

gotchas!

```{r}
1:3 %>% as.list %>% has_element(2)
1:3 %>% has_element(2)
```

instead use

```{r}
1:3 %>% as.list %>% has_element(2L)
1:3 %>% has_element(2L)

```

### invoke

Set of functions facilitating invokation / execution of combinations of functions and parameters

```{r "purrr::invoke"}
# Invoke a function with a list of arguments
invoke(runif, list(n = 10))
# Invoke a function with named arguments
invoke(runif, n = 10)

list("01a", "01b") %>%
  invoke(paste, ., sep = ".")
```

invoke map

```{r "purrr::invoke_map"}
# Invoke a list of functions, each with different arguments
invoke_map(list(runif, rnorm), list(list(n = 10), list(n = 5)))
# Or with the same inputs:
invoke_map(list(runif, rnorm), list(list(n = 5)))
invoke_map(list(runif, rnorm), n = 5)
# Or the same function with different inputs:
invoke_map("runif", list(list(n = 5), list(n = 10)))

# Or as a pipeline
list(m1 = mean, m2 = median) %>% invoke_map(x = rcauchy(100))
list(m1 = mean, m2 = median) %>% invoke_map_dbl(x = rcauchy(100))
```

### lift

composition helper. helps with sending parameters to functions.

- `d`: dots `...`
- `l`: list
- `v`: vector

`lift_dl` means lift dots to list.

```{r "purrr::lift functions"}
### Lifting from ... to list(...) or c(...)
x <- list(x = c(1:100, NA, 1000), na.rm = TRUE, trim = 0.9)
lift_dl(mean)(x)
lift(mean)(x) # lift is alias for lift_dl

# Or in a pipe:
mean %>% lift_dl %>% invoke(x)

# Default arguments can also be specified directly in lift_dl()
list(c(1:100, NA, 1000)) %>% lift_dl(mean, na.rm = TRUE)()

# lift_dl() and lift_ld() are inverse of each other.
# Here we transform sum() so that it takes a list
fun <- sum %>% lift_dl()
fun(list(3, NA, 4, na.rm = TRUE))
# Now we transform it back to a variadic function
fun2 <- fun %>% lift_ld()
fun2(3, NA, 4, na.rm = TRUE)
```

### combinations

read the `cross` help for more complex situations.

```{r "purrr::cross"}
dta <-
  list(id = c("John", "Jane"),
       greeting = c("Hello.", "Bonjour."),
       sep = c("! ", "... ")
       )

combinations <- dta %>% cross()
combinations %>% str
combinations %>% map(lift(paste))
```

### detect

`find`/`detect` in ruby. returns FIRST result based on some data and a predicate.

works for vectors or lists.

```{r "purrr::detect"}
is_even <- function(x) x %% 2 == 0

3:10 %>% detect(is_even)
3:10 %>% detect_index(is_even)

3:10 %>% detect(is_even, .right = TRUE)
3:10 %>% detect_index(is_even, .right = TRUE)

l1 <-
  list( l11 = list(a = 1, b = 2, mes = "1st item")
       ,l12 = list(a = 3, b = 4, mes = "2nd item"))
l1 %>% detect(~ .$b == 4)
```

### every / some

`all`/`any` in Ruby. predicate functions.

```{r "purrr::every"}
x <- list(0, 1, TRUE)
x %>% every(identity)
x %>% some(identity) # returns TRUE on the last predicate check

y <- list(0:10, 5.5)
y %>% every(is.numeric)
y %>% every(is.integer)
```

### flatten

similar to `unlist` but only flattens one layer at a time.

```{r "purrr::flatten"}
(x <- list( rerun(2, sample(4))
           ,rerun(2, sample(4))))
x %>% flatten()
rerun(2, sample(4)) %>% flatten_int()
```

### negate

negate a predicate function

```{r "purrr::negate"}
set.seed(12)
x <- transpose(list(x = 1:10, y = rbernoulli(10)))
x %>% keep("y") %>% length() # keep the list entries with y being TRUE
x %>% keep(negate("y")) %>% length() # same as above but y being FALSE
```

### null-default

inspired from ruby's `||`

```{r "purrr::null-default"}
1 %||% 2
NULL %||% 2
NA %||% 2
integer() %||% 2
```

