## testthat

### Overview

Section for all things hadley's testing package `testthat`.

### tests

No news is good news. successful test silently return the first argument.

- `expect_identical` tests with `identical`
- `expect_equal` tests with `all.equal`
- `expect_equivalent` tests with `all.equal` and `check.attributes = FALSE`
- with multiple assertions in the same `it`-block, all assertions are tested and any false assertion gets reported

```{r "testthat tests"}
# library(testthat)

# testing equality
a <- 10
testthat::expect_equal(a, 10)
testthat::expect_equal(10L, 10)
# Use expect_equal() when testing for numeric equality
testthat::expect_equal(sqrt(2) ^ 2, 2)

# testing logical return
testthat::expect_true(2 == 2)
testthat::expect_false(2 != 2)

# greater/less than tests
a <- 9
testthat::expect_lt(a, 10)
testthat::expect_lte(a, 9)
testthat::expect_gt(a, 8)
testthat::expect_gte(a, 9)

# regex match test
testthat::expect_match("Testing is fun", "f.n")

# type and class tests
testthat::expect_null(NULL)
testthat::expect_type(1, "double") # type test
testthat::expect_is(1, "numeric")  # class test

testthat::expect_length(1:10, 10)

# output expectations
testthat::expect_output(str(mtcars), "32 obs")
testthat::expect_output(str(mtcars), "11 variables")
testthat::expect_output(str(mtcars), "11 VARIABLES", ignore.case = TRUE) # extra arguments are sent to grepl
# -- Messages
f <- function(x) {
  if (x < 0) message("*x* is already negative")
  -x
}
testthat::expect_message(f(-1))
testthat::expect_message(f(-1), "already negative")
testthat::expect_message(f(1), NA)
# -- Warnings
f <- function(x) {
  if (x < 0) warning("*x* is already negative")
  -x
}
testthat::expect_warning(f(-1))
testthat::expect_warning(f(-1), "already negative")
testthat::expect_warning(f(1), NA)
## -- Errors
f <- function() stop("My error!")
testthat::expect_error(f())
testthat::expect_error(f(), "My error!")
## -- Silence
# doesn't check silent/invisible return, but rather if there's messages/warnings/errors being output
testthat::expect_silent("123")

# failure expectation
testthat::expect_failure(testthat::expect_identical(1, 1L))
```

### stubs

implemented with `testthat::with_mock`

```{r "with_mock; NOTE: Can't mock functions in base packages anymore", eval = FALSE}
testthat::with_mock(
   all.equal = function(x, y, ...) TRUE
  ,testthat::expect_equal(2 * 3, 4) # silently returns the return of the first argument
  ,.env = "base"
) # return the output of the last expression

testthat::with_mock(
   `base::identical` = function(x, y, ...) TRUE
  ,testthat::expect_identical(3, 5)
  ,testthat::expect_identical(TRUE, 6)
)
```

